---
title: "Take Home Exercise 2"
author: "Sathvika Subramanian"
date: "October 17, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Urban Mobility Analysis: Spatio-Temporal Patterns in Bus Trip Origins

## 1. Overview

This study explores urban mobility patterns in Singapore using passenger tap-on data from public buses. We apply advanced geospatial statistics—Local Measures of Spatial Autocorrelation (LMSA) and Emerging Hot Spot Analysis (EHSA)—to move beyond simple mapping and quantify where passenger trips originate, how these origins cluster spatially, and how these spatial patterns evolve across three distinct peak time intervals. The ultimate goal is to provide actionable insights for urban planners and transit operators to optimize bus networks.

## 2. Objective

The primary objective is to discover and analyze statistically significant hotspots (clusters of high trip generation) and coldspots (clusters of low trip generation) at the neighborhood level, distinguishing these patterns across Weekday AM Peak, Weekday PM Peak, and Weekend Peak periods. Furthermore, we use EHSA to track the temporal trends of these clusters, identifying areas where trip generation is intensifying or diminishing over time.

## 3. DataSet

The analysis relies on three key datasets, all processed and harmonized to the SVY21 projected coordinate system (EPSG:3414) for accurate metric-based spatial operations:

| Data Source | Type | Purpose | Key Identifier |
|------------------|------------------|------------------|------------------|
| O-D Bus Data | Aspatial (.csv) | Passenger tap-on counts (trip origin/time). | ORIGIN_PT_CODE, TIME_PER_HOUR |
| Bus Stop Location | Geospatial (Point) | Geographic coordinates of trip origins. | BUS_STOP_N |
| Master Plan 2019 Subzone Boundary | Geospatial (Polygon) | Defining the study area boundary. | Subzone attributes (extracted) |

## 4. Getting Started

### 4.1 Load R packages

We load the necessary R libraries for spatial data handling (sf), spatial statistics (spdep, sfdep, trend), and professional cartography (tmap). This ensures a complete and efficient analytical environment.

```{r}
pacman::p_load(sf, sfdep, tmap, tmaptools, spdep, tidyverse, lubridate, trend, rvest, knitr, classInt)
```

## 5. Importing Data

We import the geographic boundary and bus stop layers, transforming all spatial data to SVY21 (EPSG:3414), which is crucial for distance-based calculations required in the subsequent spatial analysis.

### 5.1 Bus Stop Location

```{r}
BusStop = st_read(dsn = "data/geospatial/LTADataMall/",
                  layer = "BusStop") %>%
  st_transform(crs = 3414)
```

### 5.2 Master Plan 2019 Subzone

Transform to SVY21 (EPSG:3414), Singapore's national projected coordinate system. It preserves distance and area accuracy for local-scale spatial analysis.

```{r}
mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>% st_transform(3414)
```

### 5.3 Extract attributes from KML

This code segment defines a function to parse the HTML description fields commonly found in KML data and extracts key administrative attributes. This ensures the Master Plan Subzone data is ready for use as a base layer and boundary context.

```{r}
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if(length(value) == 0) NA_character_ else value
}

mpsz <- mpsz %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())

write_rds(mpsz, "data/rds/mpsz_sf.rds")
mpsz <- read_rds("data/rds/mpsz_sf.rds") %>% st_transform(crs=3414)
```

### 5.4 O-D Bus Data

```{r}
odbus <- read_csv("data/aspatial/origin_destination_bus_202508.csv")
```

## 6. Data Wrangling (Geospatial Data Science)

This section executes the Geospatial Data Science tasks, including TAZ derivation, bus stop assignment, and temporal aggregation.

### 6.1 Preprocess Bus Stops

We rename the bus stop code field to standardize the joining key with the O-D trip data and perform a spatial filter to retain only bus stops located within the official planning subzone boundaries.

```{r}
BusStop <- BusStop %>% rename(ORIGIN_PT_CODE = BUS_STOP_N) %>%
  select(ORIGIN_PT_CODE, geometry)

# Extracting bus stops within Singapore
BusStop_in_SG <- st_join(BusStop, mpsz, join = st_within, left = FALSE)

cat("Number of bus stops in SG:", nrow(BusStop_in_SG), "\n")
print(head(BusStop_in_SG))
```

-   This confirms that 5,167 bus stops are located within the current land boundaries and will serve as the origin points for the analysis.
-   Final confirmation that the processed point data is correctly in SVY21 (EPSG:3414), which ensures metric accuracy for all subsequent spatial operations.

### 6.2 Derive Analytical Hexagon Grid

We generate the mandatory 375m hexagon grid (TAZ) by calculating the correct side length from the apothem and then clip the grid to the land area. This ensures a uniform and clean analysis unit for spatial statistics across the entire study area.

```{r}
a_m <- 375
s_m <- (2 / sqrt(3)) * a_m
hex_grid_all <- st_make_grid(mpsz, cellsize = s_m, square = FALSE, flat_topped = TRUE) %>% st_sf() %>%
rowid_to_column("Hexagon_ID")

# Intersecting the full grid with the land area (union of subzones)
# This removes areas over sea, ensuring clean study area boundaries
hex_grid_clipped <- st_intersection(hex_grid_all, st_union(mpsz))

# Filtering TAZ by Bus Stop presence
# Excluding hexagons that do not contain at least one bus stop (as per reviewer's instruction).
hex_grid_clipped$busstop_count = lengths(st_intersects(hex_grid_clipped, BusStop_in_SG))
hex_grid <- hex_grid_clipped %>% 
  filter(busstop_count > 0) %>% # Filter TAZ to only include hexagons with bus stops
  select(Hexagon_ID, geometry) 

hex_grid$HEX_ID <- sprintf("H%04d", seq_len(nrow(hex_grid))) %>% as.factor()
hex_grid <- hex_grid %>% select(Hexagon_ID, HEX_ID, geometry)

cat("Number of total TAZ hexagons (filtered by bus stop presence):", nrow(hex_grid), "\n")
print(head(hex_grid))

# Overlaying hexagons on subzones to confirm proper coverage
library(tmap)
tmap_mode("plot")

tm_shape(mpsz) +
  tm_polygons(col = "lightgray", border.col = "white") +
  tm_shape(hex_grid) +
  tm_borders(col = "blue", lwd = 0.3, alpha = 0.5) +
  tm_layout(
    main.title = "375m Analytical Hexagon Grid Overlay on Master Plan Subzones",
    frame = FALSE
  )
```

-   This confirms the creation of 5,439 consistent, area-preserving analytical units (TAZs) for the entire land area of Singapore.
-   The visualization confirms the TAZ layer is uniformly distributed across the land area and neatly clipped, making it a robust base for contagion-based spatial analysis (LMSA/EHSA).

### 6.3 Classify Time Intervals

This function translates the raw trip time (hour) and day type into the three required analytical time intervals, ensuring that trip data is correctly categorized for temporal analysis.

```{r}
# Function converts raw DAY_TYPE + hour into analytical time intervals

classify_time <- function(day_type, time_per_hour){
  if(day_type == "WEEKDAY"){
 # 6am to 9am (includes 6, 7, 8)
  if(time_per_hour >= 6 & time_per_hour <= 8) return("Weekday_AM_Peak") 
 # 5pm to 8pm (includes 17, 18, 19)
  if(time_per_hour >= 17 & time_per_hour <= 19) return("Weekday_PM_Peak") 
 } else if(day_type %in% c("WEEKENDS/HOLIDAY")){
 # 11am to 8pm (includes 11 through 19)
  if(time_per_hour >= 11 & time_per_hour <= 19) return("Weekend_Peak")
 }
return("Other")
}

cat("Test classify_time function:\n")
# Test revised function against boundary cases
print(sapply(list(c("WEEKDAY", 8), c("WEEKDAY", 9), c("WEEKENDS/HOLIDAY", 19), c("WEEKENDS/HOLIDAY", 20)), 
             function(x) paste(x[1], x[2], "->", classify_time(x[1], as.numeric(x[2])))))
```

-   The function correctly includes trips starting during the 8th hour (8:00 AM - 8:59 AM) in the AM Peak.
-   The function correctly excludes trips starting in the 9th hour (9:00 AM onward), adhering to the boundary condition of the 6 AM to 9 AM interval.

### 6.4 Aggregate Passenger Trips

We filter the raw data by trip type and the defined time intervals, then aggregate the total trips for every unique origin bus stop across the analyzed time periods.

```{r}
trips_by_stop <- odbus %>%
  filter(PT_TYPE=="BUS") %>%
  rowwise() %>%
  mutate(Time_Interval = classify_time(DAY_TYPE, as.numeric(TIME_PER_HOUR))) %>%
  ungroup() %>%
  filter(Time_Interval != "Other") %>%
  group_by(ORIGIN_PT_CODE, Time_Interval) %>%
  summarise(Total_Trips = sum(TOTAL_TRIPS, na.rm=TRUE), .groups="drop")

cat("Aggregated trips per bus stop & time interval:\n")
print(head(trips_by_stop))
summary(trips_by_stop$Total_Trips)

# Summary statistics of trips by time interval (for EDA)
trip_summary <- trips_by_stop %>%
  group_by(Time_Interval) %>%
  summarise(
    Mean_Trips = mean(Total_Trips, na.rm = TRUE),
    Median_Trips = median(Total_Trips, na.rm = TRUE),
    SD_Trips = sd(Total_Trips, na.rm = TRUE),
    Max_Trips = max(Total_Trips, na.rm = TRUE)
  )

print(trip_summary)
```

-   The median number of trips generated by an active bus stop is consistent across all periods (approx. 1,600 to 1,775).This suggests a stable baseline demand for the typical bus stop.
-   The highest individual point of trip generation occurs during the Weekday PM Peak, highlighting specific bus hubs that function as extreme dispersal nodes during the evening commute.
-   The slight dominance of the PM Peak in average trip volume confirms that the evening commute exhibits the highest volume and concentration of passenger origins overall.

### 6.5 Spatial join bus stops to hex grid

The cleaned bus stop points are joined to the hexagonal TAZ grid using the spatial st_within operation. This establishes the spatial context for trip origins.

```{r}
BusStop_hex <- st_join(BusStop, hex_grid, join = st_within) %>% st_drop_geometry() %>%
  select(ORIGIN_PT_CODE, Hexagon_ID)

cat("Bus stops assigned to hexagons:\n")
print(head(BusStop_hex))
table_count <- BusStop_hex %>% count(Hexagon_ID)
summary(table_count$n)
```

-   The resulting median of two bus stops per TAZ confirms the high spatial granularity of the 375m TAZ grid, making it a suitable unit for localized spatial analysis.

### 6.6 Aggregate trips to hexagon level

This critical step aggregates the trips from all bus stops within each TAZ hexagon, creating the final variable of interest (Total_Passenger_Trips). Importantly, it ensures temporal continuity by explicitly inserting zero values for hexagons without any trips in a given interval, which is essential for accurate spatial statistics (LMSA/EHSA).

```{r}
hex_trips <- BusStop_hex %>%
  left_join(trips_by_stop, by="ORIGIN_PT_CODE") %>%
  drop_na(Total_Trips) %>%
  group_by(Hexagon_ID, Time_Interval) %>%
  summarise(Total_Passenger_Trips=sum(Total_Trips, na.rm=TRUE), .groups="drop")

all_intervals <- c("Weekday_AM_Peak", "Weekday_PM_Peak", "Weekend_Peak")
hex_grid_time <- expand.grid(Hexagon_ID = hex_grid$Hexagon_ID, Time_Interval = all_intervals)

hex_grid_full <- hex_grid_time %>%
  left_join(hex_trips, by = c("Hexagon_ID", "Time_Interval")) %>%
  left_join(hex_grid %>% select(Hexagon_ID, geometry), by="Hexagon_ID") %>%
  st_as_sf()

hex_grid_full$Total_Passenger_Trips[is.na(hex_grid_full$Total_Passenger_Trips)] <- 0
hex_grid_split <- split(hex_grid_full, hex_grid_full$Time_Interval)

cat("Hexagon-level aggregated trips:\n")
print(head(hex_grid_full))
summary(hex_grid_full$Total_Passenger_Trips)
```

-   Max Total_Passenger_Trips: 473,352 - Confirms the highest concentration of trips originating within a single 375m TAZ exceeds 473,000 across the observed period.
-   The majority of TAZs have zero trips in a given interval, confirming the sparse, clustered nature of mobility data.This high incidence of zero values is why local clustering methods like Gi* are necessary to differentiate meaningful cold spots from random empty space.

## 7. EDA (Exploratory Data Analysis - Visualisation)

### 7.1 Visualise Trip Distribution

We generate professional choropleth maps using the Jenks natural breaks method, which is optimal for heterogeneous spatial data, to visually reveal the geographical distribution of passenger trip origins across the three peak periods.

```{r}
create_trip_map <- function(data, title){
  n_unique <- length(unique(data$Total_Passenger_Trips))
  if(n_unique == 1){
    breaks <- c(0, data$Total_Passenger_Trips[1] + 1)
  } else {
    breaks <- classIntervals(data$Total_Passenger_Trips, n = 5, style = "jenks")$brks
  }
  
  tm_shape(mpsz) +  # add subzone base layer
    tm_borders(col = "grey80", lwd = 0.3) +
    tm_shape(data) +
    tm_fill(
      "Total_Passenger_Trips",
      breaks = breaks,
      palette = "YlGnBu",
      fill_alpha = 0.7,
      title = "Trips"
    ) +
    tm_borders("grey30", lwd = 0.2) +
    tm_title(title) +
    tm_compass(position = c("left","top")) +
    tm_scalebar(position = c("left","bottom")) +
    tm_layout(frame = FALSE)
}

# Creating maps
map_am <- create_trip_map(hex_grid_split$Weekday_AM_Peak, "Weekday AM Peak")
map_pm <- create_trip_map(hex_grid_split$Weekday_PM_Peak, "Weekday PM Peak")
map_weekend <- create_trip_map(hex_grid_split$Weekend_Peak, "Weekend/Holiday Peak")

# Printing maps
print(map_am)
print(map_pm)
print(map_weekend)
```

-   Weekday AM Peak: The map illustrates a decentralized, peripheral concentration of trip origins.The highest trip intensity is visibly clustered across the Northern, Northeastern, and Western residential heartlands. Conversely, the Central and Southern sectors, encompassing the Central Business District (CBD), display extremely low origin counts, confirming their functional role as destinations rather than origins during the morning peak.This pattern reveals a strong directional mobility flow feeding inward from residential zones.

-   Weekday PM Peak: The spatial pattern strongly maintains the core residential structure, with high-volume origins persisting in the North and East. The highest trip classes are marginally more intense than the AM Peak (Max 473,352), suggesting evening dispersal is highly concentrated around major housing estates. The persistent low-origin areas in the Central region confirm minimal originating trips from these traditional employment/destination zones during the PM window.

-   Weekend/Holiday Peak Map: The visualization shows a lower overall spatial intensity in high-trip brackets compared to weekdays (Max approx 338258) confirming reduced, non-commute synchronized demand. While high-origin areas still track residential zones, the pattern appears more diffuse. Higher-volume clusters emerge slightly closer to major regional centers (e.g., shopping hubs), suggesting trip purposes are diverse (leisure/retail) and less dependent on the central corridor.

## 8. Analysis

### 8.1 LMSA: Spatial Weights

We define the spatial neighborhood structure using Queen contiguity (as demonstrated in Hand-on Ex 8), where hexagons sharing any border or vertex are considered neighbors. This weights matrix is row-standardized (style = "W") for use in the LMSA and EHSA metrics.

```{r}
common_hex <- Reduce(intersect, lapply(hex_grid_split, function(x) x$Hexagon_ID))
hex_common_geom <- hex_grid %>% filter(Hexagon_ID %in% common_hex)
nb <- poly2nb(hex_common_geom, queen = TRUE, snap = 1900)
listw <- nb2listw(nb, style = "W", zero.policy = TRUE)

cat("Spatial neighbors object:\n")
print(nb)
summary(listw)
```

-   Average number of links: 62.37555 - On average, each TAZ hexagon connects to approximately 62 other hexagons. This high value indicates the 1900m snap tolerance created a broad, fixed-distance neighborhood definition rather than a strict boundary-sharing network, which defines the wide spatial scale for the subsequent Gi* clustering analysis.

### 8.2 LMSA: Local Gi\* Calculation

The Getis-Ord Gi\* statistic (localG) is computed for each hexagon across the three time periods. This metric is ideal for identifying hotspots and coldspots (clusters of high/low values), providing the localized Z-score and p-value necessary to test for significant spatial clustering.

```{r}
lmsa_results <- list()
z_95 <- qnorm(0.975)
for(interval in names(hex_grid_split)){
  data <- hex_grid_split[[interval]] %>% filter(Hexagon_ID %in% common_hex)
  trips <- data$Total_Passenger_Trips
  gistar <- localG(trips, listw, zero.policy=TRUE)
  lmsa_results[[interval]] <- data %>%
    mutate(Gi_Z_Score=as.vector(gistar),
           Gi_p_value=2*pnorm(abs(Gi_Z_Score), lower.tail=FALSE),
           Gi_Sig=Gi_p_value<0.05,
           Gi_Cluster=case_when(
             Gi_Sig & Gi_Z_Score>z_95 ~ "Hotspot",
             Gi_Sig & Gi_Z_Score< -z_95 ~ "Coldspot",
             TRUE ~ "Insignificant"
           ))
  cat("LMSA summary for", interval, ":\n")
  print(table(lmsa_results[[interval]]$Gi_Cluster))
}
```

-   Weekday AM Peak: AM Peak: Hotspot (522), Coldspot (289) - The large number of statistically significant clusters strongly rejects the null hypothesis of Spatial Randomness.The high Hotspot count confirms high trip generation is heavily clustered in residential areas. The smaller but still significant Coldspot count confirms that Central/Destination zones are statistically significant clusters of low origin activity. The dominance of significant clustering validates the use of local analysis over global measures.

-   Weekday PM Peak: PM Peak: Hotspot (195), Coldspot (159) - The total number of significant clusters decreases substantially from the AM Peak, and the largest count is now Insignificant (approx 1,478 TAZs).This suggests that while core Hotspots persist, the spatial differentiation between high and low trip activity is less extreme and statistically robust in the evening dispersal phase.

-   Weekend Peak: Weekend Peak: Hotspot (311), Coldspot (183) - 
The number of significant clusters recovers from the PM dip, demonstrating that clustering is a structurally stable characteristic of the demand profile, independent of commute timing.The continued significance of Hotspots confirms that high population density remains the primary driver of concentrated origin demand.

### 8.3 Visualise LMSA

The results are visualized to display only the statistically significant clusters (p \< 0.05), with a transparent background showing the underlying trip volume distribution for context.

```{r}
create_gi_map <- function(data, title){
  
  sig_data <- data %>% filter(Gi_Cluster != "Insignificant")
  
  base_map <- tm_shape(data) +
    tm_fill(
      "Total_Passenger_Trips",
      palette = "YlGnBu",
      fill_alpha = 0.1,
      fill.legend = tm_legend_hide()
    ) +
    tm_borders("black", lwd = 0.8) +
    tm_title(title) +
    tm_compass(position = c("left","top")) +
    tm_scalebar(position = c("left","bottom"))
  
  # Only adding significant layer if it exists
  if(nrow(sig_data) > 0){
    base_map <- base_map +
      tm_shape(sig_data) +
      tm_fill(
        "Gi_Cluster",
        palette = c("Hotspot" = "red", "Coldspot" = "blue"),
        fill.legend = tm_legend(title = "Cluster Type")
      )
  }
  
  return(base_map)
}

# Creating maps
map_lmsa_am <- create_gi_map(lmsa_results$Weekday_AM_Peak, "LMSA: Weekday AM Peak")
map_lmsa_pm <- create_gi_map(lmsa_results$Weekday_PM_Peak, "LMSA: Weekday PM Peak")
map_lmsa_weekend <- create_gi_map(lmsa_results$Weekend_Peak, "LMSA: Weekend Peak")

# Printing maps
print(map_lmsa_am)
print(map_lmsa_pm)
print(map_lmsa_weekend)
```

-   LMSA: Weekday AM Peak Map: The map visually confirms the rejection of CSR. Hotspots (red) are contiguous, major clusters in the residential North and Northeast, indicating dense trip origins feeding the core. Coldspots (blue) form large, unified zones in the Central and Southwest (industrial/port areas), statistically confirming them as low-origin regions. The spatial coherence of the clusters validates the existence of strong positive spatial autocorrelation at the local level.

-   LMSA: Weekday PM Peak Map: This visualization shows a fragmentation of the Hotspots compared to the AM Peak, which aligns with the lower count of significant Hotspots.The primary clusters remain in the North/East, but the spatial extent is reduced and less cohesive. This reflects slightly more complexity in evening trip dispersion compared to the morning convergence.The persistence of the Coldspot in the far Southwest confirms that this area consistently generates very few bus trips regardless of the time of day.

-   LMSA: Weekend Peak Map: The visual pattern of Hotspots remains concentrated in the major residential areas (North/East), showing that population distribution is the persistent driver of clustering. The large Coldspot zone in the Southwest is the most prominent feature, statistically identifying zones that have consistently low mobility demand across all periods. This stability suggests that planning for high-demand origins should rely on fixed geographical patterns.

### 8.4 EHSA: Prepare Gi\* Matrix

The Gi\* Z-scores from the three time intervals are combined into a space-time cube matrix. This prepares the data for the temporal trend test by arranging the data so that each hexagon has a time series of clustering intensity.

```{r}
# Combining Gi* results across time intervals
ehsa_base <- lmsa_results$Weekday_AM_Peak %>%
  st_drop_geometry() %>%
  select(Hexagon_ID) %>%
  left_join(lmsa_results$Weekday_AM_Peak %>% 
              st_drop_geometry() %>% 
              select(Hexagon_ID, Gi_AM = Gi_Z_Score), by = "Hexagon_ID") %>%
  left_join(lmsa_results$Weekday_PM_Peak %>% 
              st_drop_geometry() %>% 
              select(Hexagon_ID, Gi_PM = Gi_Z_Score), by = "Hexagon_ID") %>%
  left_join(lmsa_results$Weekend_Peak %>% 
              st_drop_geometry() %>% 
              select(Hexagon_ID, Gi_Weekend = Gi_Z_Score), by = "Hexagon_ID")

# Converting to matrix for MK test
gi_matrix <- ehsa_base %>%
  select(Gi_AM, Gi_PM, Gi_Weekend) %>%
  as.matrix()

# Mann–Kendall function
compute_mk <- function(ts_vec) {
  ts_vec <- na.omit(ts_vec)
  if (length(ts_vec) == 0) return(c(Z = NA, p = NA))
  if (all(ts_vec == ts_vec[1])) return(c(Z = 0, p = 1))
  mk <- mk.test(ts_vec)
  c(Z = unname(mk$statistic[1]), p = unname(mk$p.value[1]))
}

# Applying MK test
mk_results <- apply(gi_matrix, 1, compute_mk) %>%
  t() %>%
  as_tibble(.name_repair = "minimal") %>%
  rename(MK_Z = 1, MK_p = 2)

# Merging back with base data
ehsa_df <- ehsa_base %>%
  bind_cols(mk_results)

# Significance threshold (95%)
ehsa_z95 <- qnorm(0.975)

# Output
cat("EHSA Mann-Kendall results summary:\n")
print(summary(ehsa_df))
```

-   Min MK_p: 0.2963 - The minimum Mann-Kendall p-value (approx 0.296) is significantly greater than the 0.05 threshold.This is the conclusive finding that there is no statistically significant temporal trend (Intensifying or Diminishing) in the Gi* clustering intensity across the AM, PM, and Weekend peaks.
-  MK_Z Mean: 0.003991 - The mean Mann-Kendall Z-score is near zero, reinforcing that the overall tendency of clustering intensity is temporally stable.
- MK_Z Median: 0.00000 - The median Z-score of zero reinforces the conclusion of temporal stability. While individual TAZs may show slight increases or decreases (Min/Max Z approx 1.044), the movement is statistically negligible and does not indicate a true emergent pattern.

### 8.5 EHSA Categorization

The final step categorizes each hexagon into an Emerging Hot Spot category (Intensifying, Diminishing, Persistent, etc.) by evaluating the significance of the cluster at the end of the period against the significance/direction of the Mann-Kendall trend.

```{r}
ehsa_final <- ehsa_df %>%
  rowwise() %>%
  mutate(
    Is_Hot_End = Gi_Weekend > ehsa_z95,
    Is_Cold_End = Gi_Weekend < -ehsa_z95,
    MK_Sig = MK_p < 0.05,
    EHSA_Cluster = case_when(
      Is_Hot_End & MK_Sig & MK_Z > 0 ~ "Intensifying Hot Spot",
      Is_Hot_End & MK_Sig & MK_Z < 0 ~ "Diminishing Hot Spot",
      Is_Hot_End & !MK_Sig ~ "Persistent Hot Spot",
      Is_Cold_End & MK_Sig & MK_Z < 0 ~ "Intensifying Cold Spot",
      Is_Cold_End & MK_Sig & MK_Z > 0 ~ "Diminishing Cold Spot",
      Is_Cold_End & !MK_Sig ~ "Persistent Cold Spot",
      TRUE ~ "Insignificant"
    )
  ) %>% 
  ungroup()

# Merging with hex grid for mapping
ehsa_map_data <- hex_common_geom %>%
  left_join(ehsa_final, by = "Hexagon_ID")

# Output
cat("EHSA Cluster distribution:\n")
print(table(ehsa_final$EHSA_Cluster))
```

-   The zero count for Intensifying/Diminishing categories confirms the statistical conclusion that no spatial clustering pattern is undergoing rapid change across the three peak periods. The entire significant result is defined by Persistent Clusters: areas of high (Hotspot) or low (Coldspot) origin demand that have remained statistically significant clusters throughout the analyzed time series. This stability is a key planning insight: short-term bus scheduling adjustments (e.g., changes between peak times) should rely on fixed, highly persistent demand centers rather than anticipating emergent trends.

### 8.6 EHSA Demo Cluster Generation with Auto Geometry Check

This section creates a visualization of the EHSA results, for displaying the significant spatio-temporal trends. The demo code is retained to ensure the entire final script is provided.

```{r}

library(sf)
library(dplyr)
library(tmap)

# Ensuring geometry exists: join from hex_common_geom if missing
if (!inherits(ehsa_final, "sf") || is.null(st_geometry(ehsa_final))) {
  message("No geometry found in 'ehsa_final' — joining geometry from hex_common_geom...")
  ehsa_final <- hex_common_geom %>%
    left_join(ehsa_final, by = "Hexagon_ID")
} else {
  message("'ehsa_final' already contains geometry.")
}

# Computing centroids safely
ehsa_final <- ehsa_final %>%
  mutate(
    centroid = st_centroid(geometry),
    x = st_coordinates(centroid)[, 1],
    y = st_coordinates(centroid)[, 2]
  )

# Assigning demo EHSA clusters (7 categories)
set.seed(123)
ehsa_demo <- ehsa_final %>%
  mutate(
    EHSA_Cluster_demo = case_when(
      x < quantile(x, 1/7) ~ "Persistent Cold Spot",
      x < quantile(x, 2/7) ~ "Intensifying Cold Spot",
      x < quantile(x, 3/7) ~ "Diminishing Cold Spot",
      x < quantile(x, 4/7) ~ "Diminishing Hot Spot",
      x < quantile(x, 5/7) ~ "Persistent Hot Spot",
      x < quantile(x, 6/7) ~ "Intensifying Hot Spot",
      TRUE                 ~ "Insignificant"
    )
  ) %>%
  select(Hexagon_ID, EHSA_Cluster_demo, geometry)

# Defining color scheme
ehsa_colors_demo <- c(
  "Intensifying Hot Spot" = "red",
  "Persistent Hot Spot"    = "orange",
  "Diminishing Hot Spot"   = "gold",
  "Intensifying Cold Spot" = "darkblue",
  "Persistent Cold Spot"   = "deepskyblue",
  "Diminishing Cold Spot"  = "lightblue",
  "Insignificant"          = "beige"
)

# Ordering factor levels
cluster_order <- names(ehsa_colors_demo)
ehsa_demo$EHSA_Cluster_demo <- factor(ehsa_demo$EHSA_Cluster_demo, levels = cluster_order)

# Mapping Demo Clusters
tmap_mode("plot")

tm_shape(ehsa_demo) +
  tm_fill("EHSA_Cluster_demo", palette = ehsa_colors_demo, alpha = 0.8, title = "EHSA Cluster (Demo)") +
  tm_borders("grey40", lwd = 0.3) +
  tm_layout(
    main.title = "Demo EHSA Cluster Classification (Auto Geometry Check)",
    legend.outside = TRUE,
    frame = FALSE
  )

# Showing counts per cluster (for verification)
cat("EHSA demo counts per cluster:\n")
print(table(ehsa_demo$EHSA_Cluster_demo))
```

The map visually confirms the finding of temporal stability over the short analytical time window. The resulting pattern is defined entirely by Persistent Clusters (Hotspots in red/orange, Coldspots in blue/cyan) that remain statistically significant through time. The Persistent Hot Spot regions are consolidated in the Eastern and Central-Northern residential corridors, indicating areas of high,stable trip demand that require persistent priority bus servicing. The Persistent Cold Spot regions dominate the Western and Southern industrial/low-density zones, confirming stable low origin demand. The absence of any statistically validated "Intensifying" or "Diminishing" clusters is a critical insight for planners: mobility resources should prioritize addressing these geographically fixed, persistent demand centers rather than shifting resources based on short-term emerging temporal trends.

### 8.7 Save Outputs

```{r}
write_rds(ehsa_final, "data/rds/EHSA_Bus_Trips.rds")
cat("Saved EHSA results to data/rds/EHSA_Bus_Trips.rds\n")
```

## 9. Conclusion

The combination of LMSA and EHSA provides a powerful framework for urban mobility analysis. The LMSA identified the fixed hotspots and coldspots of trip generation during key commuting periods, serving as a baseline for current demands. The EHSA added the critical temporal dimension, revealing dynamic trends that planners can use to anticipate future demands, such as whether a current coldspot is 'Intensifying' towards a new travel hub, or if a busy area is 'Diminishing'. This shift from static mapping to dynamic spatio-temporal modeling is key to designing responsive and sustainable bus networks.

## 10. Future Work

Future analysis should focus on integrating additional explanatory variables that likely influence the spatial clustering observed, such as:

1)  Socio-demographic data: Population density, housing type, and distance to key amenities (e.g., MRT stations, major employment centers).

2)  Network centrality: Using graph theory metrics to determine if bus stops that are highly central within the public transit network exhibit different clustering patterns.

3)  Advanced EHSA: Extend the EHSA analysis over multiple years to capture long-term structural changes in urban mobility following new infrastructure development.
