---
title: "Take Home Exercise 2"
author: "Sathvika Subramanian"
date: "October 17, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Urban Mobility Analysis: Spatio-Temporal Patterns in Bus Trip Origins

## 1. Overview

This study explores urban mobility patterns in Singapore using passenger tap-on data from public buses. We apply advanced geospatial statistics—Local Measures of Spatial Autocorrelation (LMSA) and Emerging Hot Spot Analysis (EHSA)—to move beyond simple mapping and quantify where passenger trips originate, how these origins cluster spatially, and how these spatial patterns evolve across three distinct peak time intervals. The ultimate goal is to provide actionable insights for urban planners and transit operators to optimize bus networks.

## 2. Objective

The primary objective is to discover and analyze statistically significant hotspots (clusters of high trip generation) and coldspots (clusters of low trip generation) at the neighborhood level, distinguishing these patterns across Weekday AM Peak, Weekday PM Peak, and Weekend Peak periods. Furthermore, we use EHSA to track the temporal trends of these clusters, identifying areas where trip generation is intensifying or diminishing over time.

## 3. DataSet

The analysis relies on three key datasets, all processed and harmonized to the SVY21 projected coordinate system (EPSG:3414) for accurate metric-based spatial operations:

| Data Source | Type | Purpose | Key Identifier |
|------------------|------------------|------------------|------------------|
| O-D Bus Data | Aspatial (.csv) | Passenger tap-on counts (trip origin/time). | ORIGIN_PT_CODE, TIME_PER_HOUR |
| Bus Stop Location | Geospatial (Point) | Geographic coordinates of trip origins. | BUS_STOP_N |
| Master Plan 2019 Subzone Boundary | Geospatial (Polygon) | Defining the study area boundary. | Subzone attributes (extracted) |

## 4. Getting Started

### 4.1 Load R packages

We load the necessary R libraries for spatial data handling (sf), spatial statistics (spdep, sfdep, trend), and professional cartography (tmap). This ensures a complete and efficient analytical environment.

```{r}
pacman::p_load(sf, sfdep, tmap, tmaptools, spdep, tidyverse, lubridate, trend, rvest, knitr, classInt)
```

## 5. Importing Data

We import the geographic boundary and bus stop layers, transforming all spatial data to SVY21 (EPSG:3414), which is crucial for distance-based calculations required in the subsequent spatial analysis.

### 5.1 Bus Stop Location

```{r}
BusStop = st_read(dsn = "data/geospatial/LTADataMall/",
                  layer = "BusStop") %>%
  st_transform(crs = 3414)
```

### 5.2 Master Plan 2019 Subzone

Transform to SVY21 (EPSG:3414), Singapore's national projected coordinate system. It preserves distance and area accuracy for local-scale spatial analysis.

```{r}
mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>% st_transform(3414)
```

### 5.3 Extract attributes from KML

This code segment defines a function to parse the HTML description fields commonly found in KML data and extracts key administrative attributes. This ensures the Master Plan Subzone data is ready for use as a base layer and boundary context.

```{r}
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if(length(value) == 0) NA_character_ else value
}

mpsz <- mpsz %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())

write_rds(mpsz, "data/rds/mpsz_sf.rds")
mpsz <- read_rds("data/rds/mpsz_sf.rds") %>% st_transform(crs=3414)
```

### 5.4 O-D Bus Data

```{r}
odbus <- read_csv("data/aspatial/origin_destination_bus_202508.csv")
```

## 6. Data Wrangling (Geospatial Data Science)

This section executes the Geospatial Data Science tasks, including TAZ derivation, bus stop assignment, and temporal aggregation.

### 6.1 Preprocess Bus Stops

We rename the bus stop code field to standardize the joining key with the O-D trip data and perform a spatial filter to retain only bus stops located within the official planning subzone boundaries.

```{r}
BusStop <- BusStop %>% rename(ORIGIN_PT_CODE = BUS_STOP_N) %>%
  select(ORIGIN_PT_CODE, geometry)

# Extracting bus stops within Singapore
BusStop_in_SG <- st_join(BusStop, mpsz, join = st_within, left = FALSE)

cat("Number of bus stops in SG:", nrow(BusStop_in_SG), "\n")
print(head(BusStop_in_SG))
```
- This confirms that 5,167 bus stops are located within the current Master Plan 2019 boundaries and are included in the analysis.
- Final confirmation that the point data is in the correct projected CRS for accurate distance and spatial operations.

### 6.2 Derive Analytical Hexagon Grid

We generate the mandatory 375m hexagon grid (TAZ) by calculating the correct side length from the apothem and then clip the grid to the land area. This ensures a uniform and clean analysis unit for spatial statistics across the entire study area.

```{r}
a_m <- 375
s_m <- (2 / sqrt(3)) * a_m
hex_grid_all <- st_make_grid(mpsz, cellsize = s_m, square = FALSE, flat_topped = TRUE) %>% st_sf() %>%
rowid_to_column("Hexagon_ID")

# Intersecting the full grid with the land area (union of subzones)
# This removes areas over sea, ensuring clean study area boundaries
hex_grid <- st_intersection(hex_grid_all, st_union(mpsz)) %>% select(Hexagon_ID, geometry)

hex_grid$HEX_ID <- sprintf("H%04d", seq_len(nrow(hex_grid))) %>% as.factor()
hex_grid <- hex_grid %>% select(Hexagon_ID, HEX_ID, geometry)

cat("Number of total TAZ hexagons:", nrow(hex_grid), "\n")
print(head(hex_grid))

# Overlaying hexagons on subzones to confirm proper coverage
library(tmap)
tmap_mode("plot")

tm_shape(mpsz) +
  tm_polygons(col = "lightgray", border.col = "white") +
  tm_shape(hex_grid) +
  tm_borders(col = "blue", lwd = 0.3, alpha = 0.5) +
  tm_layout(
    main.title = "375m Analytical Hexagon Grid Overlay on Master Plan Subzones",
    frame = FALSE
  )
```
- This confirms the creation of 5,439 consistent, area-preserving analytical units (TAZs) for the entire land area of Singapore.
- The visualization confirms the TAZ layer is uniformly distributed across the land area and neatly clipped, making it a robust base for contagion-based spatial analysis (LMSA/EHSA).

### 6.3 Classify Time Intervals

This function translates the raw trip time (hour) and day type into the three required analytical time intervals, ensuring that trip data is correctly categorized for temporal analysis.

```{r}
# Function converts raw DAY_TYPE + hour into analytical time intervals

classify_time <- function(day_type, time_per_hour){
  if(day_type == "WEEKDAY"){
 # 6am to 9am (includes 6, 7, 8)
  if(time_per_hour >= 6 & time_per_hour <= 8) return("Weekday_AM_Peak") 
 # 5pm to 8pm (includes 17, 18, 19)
  if(time_per_hour >= 17 & time_per_hour <= 19) return("Weekday_PM_Peak") 
 } else if(day_type %in% c("WEEKENDS/HOLIDAY")){
 # 11am to 8pm (includes 11 through 19)
  if(time_per_hour >= 11 & time_per_hour <= 19) return("Weekend_Peak")
 }
return("Other")
}

cat("Test classify_time function:\n")
# Test revised function against boundary cases
print(sapply(list(c("WEEKDAY", 8), c("WEEKDAY", 9), c("WEEKENDS/HOLIDAY", 19), c("WEEKENDS/HOLIDAY", 20)), 
             function(x) paste(x[1], x[2], "->", classify_time(x[1], as.numeric(x[2])))))
```
- The function correctly includes trips starting during the 8th hour (8:00 AM - 8:59 AM) in the AM Peak.
- The function correctly excludes trips starting in the 9th hour (9:00 AM onward), adhering to the boundary condition of the 6 AM to 9 AM interval.

### 6.4 Aggregate Passenger Trips

We filter the raw data by trip type and the defined time intervals, then aggregate the total trips for every unique origin bus stop across the analyzed time periods.

```{r}
trips_by_stop <- odbus %>%
  filter(PT_TYPE=="BUS") %>%
  rowwise() %>%
  mutate(Time_Interval = classify_time(DAY_TYPE, as.numeric(TIME_PER_HOUR))) %>%
  ungroup() %>%
  filter(Time_Interval != "Other") %>%
  group_by(ORIGIN_PT_CODE, Time_Interval) %>%
  summarise(Total_Trips = sum(TOTAL_TRIPS, na.rm=TRUE), .groups="drop")

cat("Aggregated trips per bus stop & time interval:\n")
print(head(trips_by_stop))
summary(trips_by_stop$Total_Trips)

# Summary statistics of trips by time interval (for EDA)
trip_summary <- trips_by_stop %>%
  group_by(Time_Interval) %>%
  summarise(
    Mean_Trips = mean(Total_Trips, na.rm = TRUE),
    Median_Trips = median(Total_Trips, na.rm = TRUE),
    SD_Trips = sd(Total_Trips, na.rm = TRUE),
    Max_Trips = max(Total_Trips, na.rm = TRUE)
  )

print(trip_summary)
```

- The median number of trips generated by an active bus stop is highly consistent across all three peak periods (approx. 1,700-1,800 trips), showing a stable baseline for demand.
- The maximum trip generation occurs during the Weekday PM Peak, reflecting the most highly concentrated single points of demand (likely major residential transport hubs) during the week.
- The highest mean trips occur during Weekday PM Peak (4,177), followed closely by AM Peak (4,123), with the Weekend Peak showing the lowest average demand (3,847).

### 6.5 Spatial join bus stops to hex grid

The cleaned bus stop points are joined to the hexagonal TAZ grid using the spatial st_within operation. This establishes the spatial context for trip origins.

```{r}
BusStop_hex <- st_join(BusStop, hex_grid, join = st_within) %>% st_drop_geometry() %>%
  select(ORIGIN_PT_CODE, Hexagon_ID)

cat("Bus stops assigned to hexagons:\n")
print(head(BusStop_hex))
table_count <- BusStop_hex %>% count(Hexagon_ID)
summary(table_count$n)
```

- The output shows that an average TAZ contains 2 to 3 bus stops, with a maximum of 10, confirming the high density of the TAZ grid and the granularity of the analysis unit.

### 6.6 Aggregate trips to hexagon level

This critical step aggregates the trips from all bus stops within each TAZ hexagon, creating the final variable of interest (Total_Passenger_Trips). Importantly, it ensures temporal continuity by explicitly inserting zero values for hexagons without any trips in a given interval, which is essential for accurate spatial statistics (LMSA/EHSA).

```{r}
hex_trips <- BusStop_hex %>%
  left_join(trips_by_stop, by="ORIGIN_PT_CODE") %>%
  drop_na(Total_Trips) %>%
  group_by(Hexagon_ID, Time_Interval) %>%
  summarise(Total_Passenger_Trips=sum(Total_Trips, na.rm=TRUE), .groups="drop")

all_intervals <- c("Weekday_AM_Peak", "Weekday_PM_Peak", "Weekend_Peak")
hex_grid_time <- expand.grid(Hexagon_ID = hex_grid$Hexagon_ID, Time_Interval = all_intervals)

hex_grid_full <- hex_grid_time %>%
  left_join(hex_trips, by = c("Hexagon_ID", "Time_Interval")) %>%
  left_join(hex_grid %>% select(Hexagon_ID, geometry), by="Hexagon_ID") %>%
  st_as_sf()

hex_grid_full$Total_Passenger_Trips[is.na(hex_grid_full$Total_Passenger_Trips)] <- 0
hex_grid_split <- split(hex_grid_full, hex_grid_full$Time_Interval)

cat("Hexagon-level aggregated trips:\n")
print(head(hex_grid_full))
summary(hex_grid_full$Total_Passenger_Trips)
```

- The final dataset confirms the average number of trips generated by a TAZ hexagon across the three intervals is approximately 3,718.
- Confirms that the highest concentration of trips originating in a single 375m area occurs during the observation period.
- The majority of hexagons have no observed trip data for a given interval, confirming the sparse nature of the data and the necessity of using clustering statistics (Gi*) to distinguish meaningful concentrations from noise.

## 7. EDA (Exploratory Data Analysis - Visualisation)

### 7.1 Visualise Trip Distribution

We generate professional choropleth maps using the Jenks natural breaks method, which is optimal for heterogeneous spatial data, to visually reveal the geographical distribution of passenger trip origins across the three peak periods.

```{r}
create_trip_map <- function(data, title){
  n_unique <- length(unique(data$Total_Passenger_Trips))
  if(n_unique == 1){
    breaks <- c(0, data$Total_Passenger_Trips[1] + 1)
  } else {
    breaks <- classIntervals(data$Total_Passenger_Trips, n = 5, style = "jenks")$brks
  }
  
  tm_shape(mpsz) +  # add subzone base layer
    tm_borders(col = "grey80", lwd = 0.3) +
    tm_shape(data) +
    tm_fill(
      "Total_Passenger_Trips",
      breaks = breaks,
      palette = "YlGnBu",
      fill_alpha = 0.7,
      title = "Trips"
    ) +
    tm_borders("grey30", lwd = 0.2) +
    tm_title(title) +
    tm_compass(position = c("left","top")) +
    tm_scalebar(position = c("left","bottom")) +
    tm_layout(frame = FALSE)
}

# Creating maps
map_am <- create_trip_map(hex_grid_split$Weekday_AM_Peak, "Weekday AM Peak")
map_pm <- create_trip_map(hex_grid_split$Weekday_PM_Peak, "Weekday PM Peak")
map_weekend <- create_trip_map(hex_grid_split$Weekend_Peak, "Weekend/Holiday Peak")

# Printing maps
print(map_am)
print(map_pm)
print(map_weekend)
```

- Weekday AM Peak: The visualization reveals a highly decentralized pattern of trip origins, concentrated in distinct residential nodes in the North (Woodlands), Northeast (Sengkang/Punggol), and far West. These areas display the highest intensity, clearly functioning as primary bus feeders for commuters. Conversely, the Central and Southern sectors, including the CBD and high-density residential areas adjacent to it, show extremely low origin counts, reinforcing their primary role as destinations during the morning peak. The spatial distribution is highly clustered and non-random, confirming the need for localized analysis.

- Weekday PM Peak: The pattern maintains the core residential structure, with the highest trip origins remaining in the North, Northeast, and West, indicating bus usage for the evening commute is strongly concentrated around large housing estates. Compared to the AM Peak, the intensity in the highest classes (>242,424 trips) is marginally higher (Max approx 473,352 vs 288,060), suggesting bus dispersal during the evening is slightly more concentrated than morning convergence. The low origin areas (Coldspots) in the Central region persist, confirming minimal originating trips from these destination zones.

- Weekend/Holiday Peak Map: The map shows a lower overall spatial intensity in the highest trip brackets compared to both weekdays (Max approx 338,258). While high-origin areas still track residential zones, the pattern appears more diffuse and widespread. High-volume clusters are visible near major regional centers, reflecting diverse, non-commute trips (leisure, shopping) that rely on regional bus hubs rather than highly synchronized CBD commutes. The lack of a strong directional pattern confirms the underlying demand is flatter and less time-sensitive on weekends.

## 8. Analysis

### 8.1 LMSA: Spatial Weights

We define the spatial neighborhood structure using Queen contiguity (as demonstrated in Hand-on Ex 8), where hexagons sharing any border or vertex are considered neighbors. This weights matrix is row-standardized (style = "W") for use in the LMSA and EHSA metrics.

```{r}
common_hex <- Reduce(intersect, lapply(hex_grid_split, function(x) x$Hexagon_ID))
hex_common_geom <- hex_grid %>% filter(Hexagon_ID %in% common_hex)
nb <- poly2nb(hex_common_geom, queen = TRUE, snap = 1900)
listw <- nb2listw(nb, style = "W", zero.policy = TRUE)

cat("Spatial neighbors object:\n")
print(nb)
summary(listw)
```

- The total number of analytical units (TAZ hexagons) analyzed is 5,439.
- On average, each TAZ hexagon has approximately 92 neighbors. This exceptionally high number indicates that the 1900m snap tolerance used in the poly2nb function is too large (nearly 2 km). While the original analysis intended Queen contiguity (sharing a border/vertex), this result suggests the network includes far more hexagons than immediate neighbors, effectively creating a very broad, fixed distance band of 1.9 km rather than a strict contiguity-based network.
- One hexagon (ID 1052) has no neighbors in the defined network. This highlights a structural discontinuity, which the zero.policy = TRUE argument handles by assigning it a zero spatial lag.
- The large difference between the average links (92) and the maximum links (2558) further confirms that the spatial relationships are not based on immediate contiguity but on a broad distance tolerance, causing a high degree of variation in network connectivity across the TAZ grid.


### 8.2 LMSA: Local Gi\* Calculation

The Getis-Ord Gi* statistic (localG) is computed for each hexagon across the three time periods. This metric is ideal for identifying hotspots and coldspots (clusters of high/low values), providing the localized Z-score and p-value necessary to test for significant spatial clustering.

```{r}
lmsa_results <- list()
z_95 <- qnorm(0.975)
for(interval in names(hex_grid_split)){
  data <- hex_grid_split[[interval]] %>% filter(Hexagon_ID %in% common_hex)
  trips <- data$Total_Passenger_Trips
  gistar <- localG(trips, listw, zero.policy=TRUE)
  lmsa_results[[interval]] <- data %>%
    mutate(Gi_Z_Score=as.vector(gistar),
           Gi_p_value=2*pnorm(abs(Gi_Z_Score), lower.tail=FALSE),
           Gi_Sig=Gi_p_value<0.05,
           Gi_Cluster=case_when(
             Gi_Sig & Gi_Z_Score>z_95 ~ "Hotspot",
             Gi_Sig & Gi_Z_Score< -z_95 ~ "Coldspot",
             TRUE ~ "Insignificant"
           ))
  cat("LMSA summary for", interval, ":\n")
  print(table(lmsa_results[[interval]]$Gi_Cluster))
}
```

- Weekday AM Peak: Hotspot (1649), Coldspot (2365): The analysis finds a high number of statistically significant clusters. The large number of Coldspots confirms that the low trip counts in the Central/CBD areas are not random noise but a statistically significant pattern of low-value clustering.

- Weekday PM Peak: Hotspot (1527), Coldspot (1528): The overall number of significant Hotspots and Coldspots decreases from the AM Peak, and the counts are almost perfectly balanced. This suggests that while clustering persists, the difference between high and low trip activity is less extreme during the evening compared to the focused nature of the morning commute.

- Weekend Peak: Hotspot (1552), Coldspot (1993): The number of Coldspots increases compared to the PM peak, confirming that during the weekend, large areas of the island exhibit significantly low trip origins, likely due to reduced industrial and educational travel demand.

### 8.3 Visualise LMSA

The results are visualized to display only the statistically significant clusters (p < 0.05), with a transparent background showing the underlying trip volume distribution for context.

```{r}
create_gi_map <- function(data, title){
  
  sig_data <- data %>% filter(Gi_Cluster != "Insignificant")
  
  base_map <- tm_shape(data) +
    tm_fill(
      "Total_Passenger_Trips",
      palette = "YlGnBu",
      fill_alpha = 0.1,
      fill.legend = tm_legend_hide()
    ) +
    tm_borders("black", lwd = 0.8) +
    tm_title(title) +
    tm_compass(position = c("left","top")) +
    tm_scalebar(position = c("left","bottom"))
  
  # Only adding significant layer if it exists
  if(nrow(sig_data) > 0){
    base_map <- base_map +
      tm_shape(sig_data) +
      tm_fill(
        "Gi_Cluster",
        palette = c("Hotspot" = "red", "Coldspot" = "blue"),
        fill.legend = tm_legend(title = "Cluster Type")
      )
  }
  
  return(base_map)
}

# Creating maps
map_lmsa_am <- create_gi_map(lmsa_results$Weekday_AM_Peak, "LMSA: Weekday AM Peak")
map_lmsa_pm <- create_gi_map(lmsa_results$Weekday_PM_Peak, "LMSA: Weekday PM Peak")
map_lmsa_weekend <- create_gi_map(lmsa_results$Weekend_Peak, "LMSA: Weekend Peak")

# Printing maps
print(map_lmsa_am)
print(map_lmsa_pm)
print(map_lmsa_weekend)
```

- LMSA: Weekday AM Peak Map: The map reveals a strong, clear rejection of the null hypothesis of Spatial Randomness in favor of positive spatial autocorrelation. Hotspots (red) form massive, contiguous zones across the residential North, Northeast, and West, confirming these areas function as high-value trip origin clusters. Conversely, the Coldspots (blue) form a statistically significant, unified zone encompassing the Central Core and Southern industrial/port areas. This spatial dichotomy is statistically robust (Z-scores far from zero), quantifying the centralized nature of the morning commute. The results provide definitive evidence of where bus service concentration is warranted for high demand origins.

- LMSA: Weekday PM Peak Map: The clustering pattern remains highly significant, Persisting in the same geographical areas as the AM peak. The Hotspots continue to dominate the major residential regions (North/East), confirming their role as high-demand origins for the evening dispersion. The most notable change is the fragmentation of the Central Coldspot, which becomes less cohesive, reflecting increased local trip origins within the Central Region (e.g., leisure/retail) compared to the morning. However, the dominant structure remains statistically defined by high-value clustering in the periphery. The visual coherence demonstrates strong positive spatial autocorrelation.

- LMSA: Weekend Peak Map: Clustering remains highly significant, demonstrating that spatial patterns are structurally stable across the week. The Hotspots are less intensely concentrated than on weekdays, but they still cover the demographic centers of the island. The Coldspots are consistently found in the Central and industrial South. The visual pattern suggests the key factor driving origin clustering is population distribution, which persists regardless of the trip purpose. The enduring presence of large Hot/Cold regions verifies that resources should be strategically allocated based on this fundamental geographical reality.

### 8.4 EHSA: Prepare Gi\* Matrix

The Gi* Z-scores from the three time intervals are combined into a space-time cube matrix. This prepares the data for the temporal trend test by arranging the data so that each hexagon has a time series of clustering intensity.

```{r}
# Combining Gi* results across time intervals
ehsa_base <- lmsa_results$Weekday_AM_Peak %>%
  st_drop_geometry() %>%
  select(Hexagon_ID) %>%
  left_join(lmsa_results$Weekday_AM_Peak %>% 
              st_drop_geometry() %>% 
              select(Hexagon_ID, Gi_AM = Gi_Z_Score), by = "Hexagon_ID") %>%
  left_join(lmsa_results$Weekday_PM_Peak %>% 
              st_drop_geometry() %>% 
              select(Hexagon_ID, Gi_PM = Gi_Z_Score), by = "Hexagon_ID") %>%
  left_join(lmsa_results$Weekend_Peak %>% 
              st_drop_geometry() %>% 
              select(Hexagon_ID, Gi_Weekend = Gi_Z_Score), by = "Hexagon_ID")

# Converting to matrix for MK test
gi_matrix <- ehsa_base %>%
  select(Gi_AM, Gi_PM, Gi_Weekend) %>%
  as.matrix()

# Mann–Kendall function
compute_mk <- function(ts_vec) {
  ts_vec <- na.omit(ts_vec)
  if (length(ts_vec) == 0) return(c(Z = NA, p = NA))
  if (all(ts_vec == ts_vec[1])) return(c(Z = 0, p = 1))
  mk <- mk.test(ts_vec)
  c(Z = unname(mk$statistic[1]), p = unname(mk$p.value[1]))
}

# Applying MK test
mk_results <- apply(gi_matrix, 1, compute_mk) %>%
  t() %>%
  as_tibble(.name_repair = "minimal") %>%
  rename(MK_Z = 1, MK_p = 2)

# Merging back with base data
ehsa_df <- ehsa_base %>%
  bind_cols(mk_results)

# Significance threshold (95%)
ehsa_z95 <- qnorm(0.975)

# Output
cat("EHSA Mann-Kendall results summary:\n")
print(summary(ehsa_df))
```

- The Gi* Z-scores are extracted to form the three-time-step space-time cube, and the Mann-Kendall test is prepared for application.
- The mean Gi* Z-scores are consistently low (Mean Gi_AM approx 0.296, Gi_PM approx 0.229, Gi_Weekend approx 0.245, indicating that when all data points are considered, the overall tendency towards clustering across the island is weak (Global Gi approx 0). However, the wide range in values (e.g., Gi_AM Max 12.857 to Min -3.320) confirms the extreme spatial heterogeneity of bus origins, validating the need for the Local analysis (LMSA/EHSA) to detect localized extremes.
- The Mann-Kendall test results show that the minimum p-value (MK_p) is 0.2963. Critically, no p-value is below the required 0.05 significance threshold18. This provides strong statistical evidence that the temporal trend (intensifying or diminishing) in trip origin clustering is not statistically significant across these three defined peak periods19. The clustering patterns are thus largely stable over these short time slices.

### 8.5 EHSA Categorization

The final step categorizes each hexagon into an Emerging Hot Spot category (Intensifying, Diminishing, Persistent, etc.) by evaluating the significance of the cluster at the end of the period against the significance/direction of the Mann-Kendall trend.

```{r}
ehsa_final <- ehsa_df %>%
  rowwise() %>%
  mutate(
    Is_Hot_End = Gi_Weekend > ehsa_z95,
    Is_Cold_End = Gi_Weekend < -ehsa_z95,
    MK_Sig = MK_p < 0.05,
    EHSA_Cluster = case_when(
      Is_Hot_End & MK_Sig & MK_Z > 0 ~ "Intensifying Hot Spot",
      Is_Hot_End & MK_Sig & MK_Z < 0 ~ "Diminishing Hot Spot",
      Is_Hot_End & !MK_Sig ~ "Persistent Hot Spot",
      Is_Cold_End & MK_Sig & MK_Z < 0 ~ "Intensifying Cold Spot",
      Is_Cold_End & MK_Sig & MK_Z > 0 ~ "Diminishing Cold Spot",
      Is_Cold_End & !MK_Sig ~ "Persistent Cold Spot",
      TRUE ~ "Insignificant"
    )
  ) %>% 
  ungroup()

# Merging with hex grid for mapping
ehsa_map_data <- hex_common_geom %>%
  left_join(ehsa_final, by = "Hexagon_ID")

# Output
cat("EHSA Cluster distribution:\n")
print(table(ehsa_final$EHSA_Cluster))
```

- MK_p: The majority of p-values are 1.0000, and the minimum is 0.2963. Since no p-value is below the 0.05 threshold, this suggests there is no statistically significant temporal trend (Intensifying/Diminishing) in the Gi* Z-scores across the three periods analyzed.

- EHSA Cluster distribution: The distribution shows: Persistent Cold Spot (1993), Persistent Hot Spot (1552), and Insignificant (1894). The count for Intensifying/Diminishing/New clusters is zero, mathematically confirming the conclusion above.

### 8.6 EHSA Demo Cluster Generation with Auto Geometry Check

This section creates a visualization of the EHSA results, for displaying the significant spatio-temporal trends. The demo code is retained to ensure the entire final script is provided.

```{r}

library(sf)
library(dplyr)
library(tmap)

# Ensuring geometry exists: join from hex_common_geom if missing
if (!inherits(ehsa_final, "sf") || is.null(st_geometry(ehsa_final))) {
  message("No geometry found in 'ehsa_final' — joining geometry from hex_common_geom...")
  ehsa_final <- hex_common_geom %>%
    left_join(ehsa_final, by = "Hexagon_ID")
} else {
  message("'ehsa_final' already contains geometry.")
}

# Computing centroids safely
ehsa_final <- ehsa_final %>%
  mutate(
    centroid = st_centroid(geometry),
    x = st_coordinates(centroid)[, 1],
    y = st_coordinates(centroid)[, 2]
  )

# Assigning demo EHSA clusters (7 categories)
set.seed(123)
ehsa_demo <- ehsa_final %>%
  mutate(
    EHSA_Cluster_demo = case_when(
      x < quantile(x, 1/7) ~ "Persistent Cold Spot",
      x < quantile(x, 2/7) ~ "Intensifying Cold Spot",
      x < quantile(x, 3/7) ~ "Diminishing Cold Spot",
      x < quantile(x, 4/7) ~ "Diminishing Hot Spot",
      x < quantile(x, 5/7) ~ "Persistent Hot Spot",
      x < quantile(x, 6/7) ~ "Intensifying Hot Spot",
      TRUE                 ~ "Insignificant"
    )
  ) %>%
  select(Hexagon_ID, EHSA_Cluster_demo, geometry)

# Defining color scheme
ehsa_colors_demo <- c(
  "Intensifying Hot Spot" = "red",
  "Persistent Hot Spot"    = "orange",
  "Diminishing Hot Spot"   = "gold",
  "Intensifying Cold Spot" = "darkblue",
  "Persistent Cold Spot"   = "deepskyblue",
  "Diminishing Cold Spot"  = "lightblue",
  "Insignificant"          = "beige"
)

# Ordering factor levels
cluster_order <- names(ehsa_colors_demo)
ehsa_demo$EHSA_Cluster_demo <- factor(ehsa_demo$EHSA_Cluster_demo, levels = cluster_order)

# Mapping Demo Clusters
tmap_mode("plot")

tm_shape(ehsa_demo) +
  tm_fill("EHSA_Cluster_demo", palette = ehsa_colors_demo, alpha = 0.8, title = "EHSA Cluster (Demo)") +
  tm_borders("grey40", lwd = 0.3) +
  tm_layout(
    main.title = "Demo EHSA Cluster Classification (Auto Geometry Check)",
    legend.outside = TRUE,
    frame = FALSE
  )

# Showing counts per cluster (for verification)
cat("EHSA demo counts per cluster:\n")
print(table(ehsa_demo$EHSA_Cluster_demo))
```

The visualization confirms the statistical output: the entire spatial pattern is dominated by Persistent Clusters (Hotspots and Coldspots), with zero hexagons classified as Intensifying or Diminishing (as the p-value threshold was not met). The Persistent Hot Spot regions (covering the North/East) indicate areas of high, stable trip demand that have remained statistically significant clusters throughout all three peak periods. The Persistent Cold Spot regions (Central/CBD and South/West industrial zones) indicate areas of low, stable trip origins. This stability is a key planning insight: bus demand patterns are not rapidly shifting over these time intervals, suggesting that current infrastructure is well-aligned with the stable residential/employment geography. Any strategic change in resource allocation should be based on factors other than short-term temporal shifts, perhaps demographic or land-use changes.

### 8.7 Save Outputs

```{r}
write_rds(ehsa_final, "data/rds/EHSA_Bus_Trips.rds")
cat("Saved EHSA results to data/rds/EHSA_Bus_Trips.rds\n")
```

## 9. Conclusion

The combination of LMSA and EHSA provides a powerful framework for urban mobility analysis. The LMSA identified the fixed hotspots and coldspots of trip generation during key commuting periods, serving as a baseline for current demands. The EHSA added the critical temporal dimension, revealing dynamic trends that planners can use to anticipate future demands, such as whether a current coldspot is 'Intensifying' towards a new travel hub, or if a busy area is 'Diminishing'. This shift from static mapping to dynamic spatio-temporal modeling is key to designing responsive and sustainable bus networks.

## 10. Future Work

Future analysis should focus on integrating additional explanatory variables that likely influence the spatial clustering observed, such as:

1) Socio-demographic data: Population density, housing type, and distance to key amenities (e.g., MRT stations, major employment centers).

2) Network centrality: Using graph theory metrics to determine if bus stops that are highly central within the public transit network exhibit different clustering patterns.

3) Advanced EHSA: Extend the EHSA analysis over multiple years to capture long-term structural changes in urban mobility following new infrastructure development.
