---
title: "In-class_Ex06"
author: "Sathvika Subramanian"
date: "Oct 4, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Take-Home Exercise 2 Kick Starter

## Loading the R package

```{r}
pacman::p_load(sf,sfdep,tmap,knitr,tidyverse,rvest)
```

## Importing Data

## Bus Stop Location

```{r}
BusStop = st_read(dsn = "data/LTADataMall/",
                  layer = "BusStop") %>%
  st_transform(crs = 3414)
```

## Master Plan 2019 Subzone

```{r}
mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
```

```{r}
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}
```

```{r}
mpsz <- mpsz %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())
```

```{r}
write_rds(mpsz, "data/rds/mpsz_sf.rds")
```

```{r}
mpsz = read_rds("data/rds/mpsz_sf.rds") %>%
  st_transform(crs=3414)
```

## O-D Bus

```{r}
odbus <- read_csv("data/LTADataMall/origin_destination_bus_202508.csv")
```


## Plot of Bus Stop and Master Plan

```{r}
tm_shape(mpsz)+
  tm_polygons() +
  tm_shape(BusStop) +
  tm_dots()
```

## Extracting Bus Stops within Singapore

```{r}
BusStop_in_SG <- st_join(
  BusStop, mpsz,
  join = st_within,
  left = FALSE)
```

## Deriving Analytical Hexagon

```{r}
hexagon <- st_make_grid(mpsz,
                        cellsize = 700,
                        what = "polygon",
                        square = FALSE) %>%
  st_sf()
```

## Checking the generated hexagons reveals that it covers all the bus stops

## Filtering hexagons with bus stops

```{r}
hexagon$n_collisions =
  lengths(st_intersects(
    hexagon, BusStop_in_SG))

hexagon <- filter(hexagon, n_collisions > 0)
```

## Assigning ids to each hexagon

```{r}
hexagon <- hexagon %>%
  select(,-n_collisions)

hexagon$HEX_ID <- sprintf(
  "H%04d", seq_len(nrow(hexagon))) %>%
  as.factor()

kable(head(hexagon))
```

## Preparing Trip Generation Data

```{r}
trips <- odbus %>%
  select(c(ORIGIN_PT_CODE, DAY_TYPE, TIME_PER_HOUR, TOTAL_TRIPS)) %>%
  rename(BUS_STOP_N = ORIGIN_PT_CODE) %>%
  rename(HOUR_OF_DAY = TIME_PER_HOUR) %>%
  rename(TRIPS = TOTAL_TRIPS)
trips$BUS_STOP_N <- as.factor(trips$BUS_STOP_N)
kable(head(trips))
```

```{r}
bs_hex <- st_intersection(
  BusStop, hexagon) %>%
  st_drop_geometry() %>%
  select(c(BUS_STOP_N, HEX_ID))
kable(head(bs_hex))
```

## Adding HEX_ID into bus trips data

```{r}
trips <- inner_join(trips, bs_hex)
kable(head(trips))
```

