---
title: "In-class_Ex04"
author: "Sathvika Subramanian"
date: "September 20, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Geographically Weighted Summary Statistics - gwModel methods

## Import libraries

```{r}
pacman::p_load(sf, ggstatsplot, tmap,tidyverse, knitr, GWmodel)
```

## Importing 'Hunan' table

```{r}
hunan_sf <- st_read(dsn = "data/geospatial", layer = "Hunan")
```

```{r}
hunan_2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

## Joining Hunan and Hunan_2012

```{r}
hunan_sf <- left_join(hunan_sf, hunan_2012) %>% 
  select(1:3, 7, 15, 16, 31, 32)
```

```{r}
hunan_sp <- hunan_sf %>%
  as_Spatial()
```

## Cross-Validation

```{r}
bw_CV <- bw.gwr(GDPPC ~ 1,
                data = hunan_sp,
                approach = "CV",
                adaptive = FALSE,
                kernel = "bisquare",
                longlat = T)
bw_CV
```

## AIC

```{r}
bw_AIC <- bw.gwr(GDPPC ~ 1,
                data = hunan_sp,
                approach = "AIC",
                adaptive = FALSE,
                kernel = "bisquare",
                longlat = T)
bw_AIC
```

## Summary Statistics

```{r}
#with CV
gwstat_CV <- gwss(data = hunan_sp,
               vars = "GDPPC",
               bw = bw_CV,
               kernel = "bisquare",
               adaptive = FALSE,
               longlat = T)

```

```{r}
#with AIC
gwstat <- gwss(data = hunan_sp,
               vars = "GDPPC",
               bw = bw_AIC,
               kernel = "bisquare",
               adaptive = FALSE,
               longlat = T)

```

## Extracting SDF data table from gwss object output from gwss()

```{r}
gwstat_df <- as.data.frame(gwstat$SDF)
```

## Combine with Hunan data

```{r}
hunan_gstat <- cbind(hunan_sf,gwstat_df)
#doing column append from gwstat_df to hunan_sf
#use cbind when joining table without unique identifier, only when row sequence are the same

```

```{r}
basemap <- tm_shape(hunan_gstat) +
  tm_polygons() +
  tm_text("NAME_3", size=0.5)

gdppc <- qtm(hunan_gstat, "GDPPC") +
  tm_layout(
    legend.position = c("left", "bottom"))
tmap_arrange(basemap, gdppc, asp=1, ncol=2)
```

## Visualising geographically weighted summary statistics

```{r}
tm_shape(hunan_gstat) +
  tm_fill("GDPPC_LM",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Distribution of geographically weighted mean",
            main.title.position = "center",
            main.title.size = 1.8,
            legend.text.size = 1.2,
            legend.height = 1.50,
            legend.width = 1.50,
            frame = TRUE)
```

## Geographically Weighted Correlation

```{r}
ggscatterstats(
  data = hunan_2012, 
  x = Agri, 
  y = GDPPC,
  xlab = "Gross Agriculture Output", ## label for the x-axis
  ylab = "GDP per capita", 
  label.var = County, 
  label.expression = Agri > 10000 & GDPPC > 50000, 
  point.label.args = list(alpha = 0.7, size = 4, color = "grey50"),
  xfill = "#CC79A7", 
  yfill = "#009E73", 
  title = "Relationship between GDP PC and Gross Agriculture Output")
```

## Geospatial analytics solution

### Determine bandwidth
```{r}
bw <- bw.gwr(GDPPC ~ GIO, 
             data = hunan_sp, 
             approach = "AICc", 
             adaptive = TRUE)
```

## Computing gwCorrelation

```{r}
gwstats <- gwss(hunan_sp, 
                vars = c("GDPPC", "GIO"), 
                bw = bw,
                kernel = "bisquare",
                adaptive = TRUE, 
                longlat = T)
```

## Extracting the result

```{r}
gwstat_df <- as.data.frame(gwstats$SDF) %>%
  select(c(12,13)) %>%
  rename(gwCorr = Corr_GDPPC.GIO,
         gwSpearman = Spearman_rho_GDPPC.GIO)
```

```{r}
hunan_Corr <- cbind(hunan_sf, gwstat_df)
```

## Visualising Local Correlation

```{r}
tm_shape(hunan_Corr) +
  tm_polygons(fill = "gwSpearman",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "local Spearman Rho")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Local Spearman Rho", 
  size = 2.0) +
  tm_layout(frame = TRUE)
```
