---
title: "Hands-on Exercise 3"
author: "Sathvika Subramanian"
date: "Sep 11, 2025"
date-modified: "last-modified"
format: html
execute:
 echo: true
 eval: true
 warning: false
 freeze: true
---

# Spatio-Temporal Point Patterns Analysis

# 1. Installing and Loading the R packages

```{r}
pacman::p_load(sf, terra, spatstat, sparr, tmap, tidyverse)
```

# 2. Importing and Preparing Study Area

## 2.1 Importing study area

```{r}
kbb <- st_read(dsn="data/data/rawdata",
               layer = "Kepulauan_Bangka_Belitung") 
```

```{r}
kbb_sf <- st_read(dsn="data/data/rawdata",
               layer = "Kepulauan_Bangka_Belitung") %>%
  st_union() %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 32748)
```

## 2.2 Converting OWIN

```{r}
kbb_owin <- as.owin(kbb_sf)
kbb_owin
```

```{r}
class(kbb_owin)
```

# 3. Importing and Preparing Forest Fire data

```{r}
fire_sf <- read_csv("data/data/rawdata/forestfires.csv") %>%
  st_as_sf(coords = c("longitude", "latitude"),
                       crs = 4326) %>%
  st_transform(crs = 32748)
```

```{r}
fire_sf <- fire_sf %>% 
  mutate(DayofYear = yday(acq_date)) %>%
  mutate(Month_num = month(acq_date)) %>%
  mutate(Month_fac = month(acq_date, 
                           label = TRUE, 
                           abbr = FALSE))
```

# 4. Visualising the Fire Points

## 4.1 Overall plot

```{r}
tm_shape(kbb_sf)+
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots()
```

## 4.2 Visuaising geographic distribution of forest fires by month

```{r}
tm_shape(kbb_sf)+
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots(size = 0.1) +
tm_facets(by="Month_fac", 
            free.coords=FALSE, 
            drop.units = TRUE)
```

# 5.Computing STKDE by Month

## 5.1 Extracting forest fires by month

```{r}
fire_month <- fire_sf %>% 
  select(Month_num)
```

## 5.2 Creating ppp

```{r}
fire_month_ppp <- as.ppp(fire_month)
fire_month_ppp
```

```{r}
summary(fire_month_ppp)
```

```{r}
any(duplicated(fire_month_ppp))
```

## 5.3 Including Owin object

```{r}
fire_month_owin <- fire_month_ppp[kbb_owin]
summary(fire_month_owin)
```

```{r}
plot(fire_month_owin)
```

## 5.4 Computing Spatio-temporal KDE

```{r}
st_kde <- spattemp.density(fire_month_owin)
summary(st_kde)
```

## 5.5 Plotting the spatio-temporal KDE object

```{r}
tims <- c(7,8,9,10,11,12)
par(mfcol=c(2,3))
for(i in tims){ 
  plot(st_kde, i, 
       override.par=FALSE, 
       fix.range=TRUE, 
       main=paste("KDE at month",i))
}
```

# 6. Computing STKDE by Day of Year

## 6.1 Creating ppp object

```{r}
fire_yday_ppp <- fire_sf %>% 
  select(DayofYear) %>%
  as.ppp()
```

## 6.2 Including Owin object

```{r}
fire_yday_owin <- fire_yday_ppp[kbb_owin]
summary(fire_yday_owin)
```

## 6.3 Computing Spatio-temporal KDE

```{r}
kde_yday <- spattemp.density(
  fire_yday_owin)
summary(kde_yday)
```

```{r}
plot(kde_yday)
```

# 7. Computing STKDE by Day of Year: Improved method

```{r}
set.seed(1234)
BOOT.spattemp(fire_yday_owin) 
```

## 7.1 Computing spatio-temporal KDE

```{r}
kde_yday <- spattemp.density(fire_yday_owin,
                             h = 9000,
                             lambda = 19)
summary(kde_yday)
```

## 7.2 Plotting the output spatio-temporal KDE

```{r}
plot(kde_yday)
```

# 8. Spatio-temporal Point Patterns Analysis: stpp methods

## 8.1 Preparing spatio-temporal point process object of stpp

```{r}
coords <- st_coordinates(fire_sf)
```

```{r}
fire_df <- data.frame(
  x = coords[, 1],  
  y = coords[, 2],
  t = fire_sf$`DayofYear`)
```

```{r}
#fire_stpp <- as.3dpoints(fire_df)
```

```{r}
#class(fire_stpp)
```

```{r}
#plot(fire_stpp)
```

## 8.2 Computing spatio-temporal k-function

```{r}
#kbb_stik <- STIKhat(fire_stpp)
```

```{r}
#plotK(kbb_stik)
```
