---
title: "Hands-on Exercise 5-1"
author: "Sathvika Subramanian"
date: "Sep 25, 2025"
date-modified: "last-modified"
format: html
execute:
 echo: true
 eval: true
 warning: false
 freeze: true
---

# Global Measures of Spatial Autocorrelation

# 1.Setting the Analytical Tools

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse)
```

# 2.Getting the Data Into R Environment

## 2.1 Import shapefile into r environment

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

## 2.2 Import csv file into r environment

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

## 2.3 Performing relational join

```{r}
hunan <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```

# 3. Visualising Regional Development Indicator

```{r}
equal <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5,
                values = "brewer.blues")) +
  tm_borders(fiil_alpha = 0.5) + 
  tm_layout(legend.position = c("left", "bottom"), 
            main.title = "Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fiil_alpha = 0.5) + 
  tm_layout(legend.position = c("left", "bottom"), 
            main.title = "Quantile classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```

# 4. Global Measures of Spatial Autocorrelation

## 4.1 Computing Contiguity Spatial Weights

```{r}
wm_q <- poly2nb(hunan, 
                queen=TRUE)
summary(wm_q)
```

## 4.2 Row-standardised weights matrix

```{r}
rswm_q <- nb2listw(wm_q, 
                   style="W", 
                   zero.policy = TRUE)
rswm_q
```

# 5. Global Measures of Spatial Autocorrelation: Moran’s I

## 5.1 Maron’s I test

```{r}
moran.test(hunan$GDPPC, 
           listw=rswm_q, 
           zero.policy = TRUE, 
           na.action=na.omit)
```

## 5.2 Computing Monte Carlo Moran’s I

```{r}
set.seed(1234)
bperm= moran.mc(hunan$GDPPC, 
                listw=rswm_q, 
                nsim=999, 
                zero.policy = TRUE, 
                na.action=na.omit)
bperm
```

## 5.3 Visualising Monte Carlo Moran’s I

```{r}
mean(bperm$res[1:999])
```

```{r}
var(bperm$res[1:999])
```

```{r}
summary(bperm$res[1:999])
```

```{r}
hist(bperm$res, 
     freq=TRUE, 
     breaks=20, 
     xlab="Simulated Moran's I")
abline(v=0, 
       col="red") 
```

```{r}
library(ggplot2)

ggplot(data.frame(res = bperm$res), aes(x = res)) +
  geom_histogram(
    bins = 20,          # equivalent to breaks = 20
    color = "black", 
    fill = "lightpink"
  ) +
  geom_vline(
    xintercept = 0, 
    color = "red", 
    linewidth = 1
  ) +
  labs(
    x = "Simulated Moran's I", 
    y = "Frequency"
  ) +
  theme_minimal()
```

# 6. Global Measures of Spatial Autocorrelation: Geary’s C

## 6.1 Geary’s C test

```{r}
geary.test(hunan$GDPPC, listw=rswm_q)
```

## 6.2 Computing Monte Carlo Geary’s C

```{r}
set.seed(1234)
bperm=geary.mc(hunan$GDPPC, 
               listw=rswm_q, 
               nsim=999)
bperm
```

## 6.3 Visualising the Monte Carlo Geary’s C

```{r}
mean(bperm$res[1:999])
```

```{r}
var(bperm$res[1:999])
```

```{r}
summary(bperm$res[1:999])
```

```{r}
hist(bperm$res, freq=TRUE, breaks=20, xlab="Simulated Geary c")
abline(v=1, col="red") 
```

# Local Measures of Spatial Autocorrelation: Local Moran’s I (LISA)

```{r}
# Computing Local Moran's I
localMI <- localmoran(hunan$GDPPC, 
                      listw = rswm_q, 
                      zero.policy = TRUE)

# Adding results back into sf object
hunan$local_I     <- localMI[, 1]
hunan$local_pval  <- localMI[, 5]

# Classifying cluster types
hunan <- hunan %>%
  mutate(
    lag_GDPPC = lag.listw(rswm_q, GDPPC),
    LISA_cluster = case_when(
      GDPPC > mean(GDPPC) & lag_GDPPC > mean(lag_GDPPC) & local_pval < 0.05 ~ "High-High",
      GDPPC < mean(GDPPC) & lag_GDPPC < mean(lag_GDPPC) & local_pval < 0.05 ~ "Low-Low",
      GDPPC > mean(GDPPC) & lag_GDPPC < mean(lag_GDPPC) & local_pval < 0.05 ~ "High-Low",
      GDPPC < mean(GDPPC) & lag_GDPPC > mean(lag_GDPPC) & local_pval < 0.05 ~ "Low-High",
      TRUE ~ "Not Significant"
    )
  )

# Plotting LISA cluster map
tm_shape(hunan) +
  tm_polygons("LISA_cluster",
              palette = "Set1",
              title = "Local Moran’s I Cluster Map") +
  tm_borders(alpha = 0.6) +
  tm_layout(main.title = "LISA (Local Moran’s I) Cluster Patterns")
```

# 7. Spatial Correlogram

## 7.1 Compute Moran’s I correlogram

```{r}
MI_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="I", 
                          style="W")
plot(MI_corr)
```

```{r}
print(MI_corr)
```

## 7.2 Compute Geary’s C correlogram and plot

```{r}
GC_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="C", 
                          style="W")
plot(GC_corr)
```

```{r}
print(GC_corr)
```
