---
title: "Hands-on Exercise 9"
author: "Sathvika Subramanian"
date: "Nov 3, 2025"
date-modified: "last-modified"
format: html
execute:
 echo: true
 eval: true
 warning: false
 freeze: true
---

# Modelling Geographical Accessibility

## 1. Installing and Loading R packages

```{r}
pacman::p_load(tmap, SpatialAcc, sf, 
               ggstatsplot, reshape2,
               tidyverse)
```

## 2. Geospatial Data Wrangling

### 2.1 Importing geospatial data

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_NO_SEA_PL")
```

```{r}
hexagons <- st_read(dsn = "data/geospatial", layer = "hexagons") 
```

```{r}
eldercare <- st_read(dsn = "data/geospatial", layer = "ELDERCARE") 
```

### 2.2 Updating CRS information

```{r}
mpsz <- st_transform(mpsz, 3414)
eldercare <- st_transform(eldercare, 3414)
hexagons <- st_transform(hexagons, 3414)
```

```{r}
st_crs(mpsz)
```

```{r}
# Checking alignment of all layers visually
tmap_mode("plot")

tm_shape(mpsz) + 
  tm_polygons(col = "lightgrey", border.col = "white", alpha = 0.5) +
tm_shape(hexagons) + 
  tm_borders(col = "blue", alpha = 0.5) +
tm_shape(eldercare) + 
  tm_dots(size = 0.1, col = "red") +
tm_layout(title = "Data Alignment Check: Subzones, Hexagons & Eldercare")
```

### 2.3 Cleaning and updating attribute fields of the geospatial data

```{r}
eldercare <- eldercare %>%
  select(fid, ADDRESSPOS) %>%
  mutate(capacity = 100)
```

```{r}
hexagons <- hexagons %>%
  select(fid) %>%
  mutate(demand = 100)
```

## 3. Apsaital Data Handling and Wrangling

### 3.1 Importing Distance Matrix

```{r}
ODMatrix <- read_csv("data/aspatial/OD_Matrix.csv", skip = 0)
```

### 3.2 Tidying distance matrix

```{r}
distmat <- ODMatrix %>%
  select(origin_id, destination_id, total_cost) %>%
  spread(destination_id, total_cost)%>%
  select(c(-c('origin_id')))
```

```{r}
distmat_km <- as.matrix(distmat/1000)
```

## 4. Modelling and Visualising Accessibility using Hansen Method

### 4.1 Computing Hansen’s accessibility

```{r}
acc_Hansen <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            d0 = 50,
                            power = 2, 
                            family = "Hansen"))
```

```{r}
colnames(acc_Hansen) <- "accHansen"
```

```{r}
acc_Hansen <- as_tibble(acc_Hansen)
```

```{r}
hexagon_Hansen <- bind_cols(hexagons, acc_Hansen)
```

```{r}
acc_Hansen <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            #d0 = 50,
                            power = 0.5, 
                            family = "Hansen"))

colnames(acc_Hansen) <- "accHansen"
acc_Hansen <- as_tibble(acc_Hansen)
hexagon_Hansen <- bind_cols(hexagons, acc_Hansen)
```

### 4.2 Visualising Hansen’s accessibility

#### 4.2.1 Extracting map extend

```{r}
mapex <- st_bbox(hexagons)
```

```{r}
tm_shape(hexagon_Hansen,
         bbox = mapex) + 
  tm_polygons(fill = "accHansen", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Accessibility to eldercare: Hansen method") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)
```

### 4.3 Statistical graphic visualisation

```{r}
hexagon_Hansen <- st_join(hexagon_Hansen, mpsz, 
                          join = st_intersects)
```

```{r}
ggplot(data=hexagon_Hansen, 
       aes(y = log(accHansen), 
           x= REGION_N)) +
  geom_boxplot() +
  geom_point(stat="summary", 
             fun.y="mean", 
             colour ="red", 
             size=2)
```

## 5. Modelling and Visualising Accessibility using KD2SFCA Method

### 5.1 Computing KD2SFCA’s accessibility

```{r}
acc_KD2SFCA <- data.frame(ac(hexagons$demand,
                            eldercare$capacity,
                            distmat_km, 
                            d0 = 5,
                            power = 2, 
                            family = "KD2SFCA"))

colnames(acc_KD2SFCA) <- "accKD2SFCA"
acc_KD2SFCA <- as_tibble(acc_KD2SFCA)
hexagon_KD2SFCA <- bind_cols(hexagons, acc_KD2SFCA)
```

### 5.2 Visualising KD2SFCA’s accessibility

```{r}
tm_shape(hexagon_KD2SFCA,
         bbox = mapex) + 
  tm_polygons(fill = "accKD2SFCA", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Accessibility to eldercare: KD2SFCA method") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)
```

### 5.3 Statistical graphic visualisation

```{r}
hexagon_KD2SFCA <- st_join(hexagon_KD2SFCA, mpsz, 
                          join = st_intersects)
```

```{r}
ggplot(data=hexagon_KD2SFCA, 
       aes(y = accKD2SFCA, 
           x= REGION_N)) +
  geom_boxplot() +
  geom_point(stat="summary", 
             fun.y="mean", 
             colour ="red", 
             size=2)
```

## 6. Modelling and Visualising Accessibility using Spatial Accessibility Measure (SAM) Method

### 6.1 Computing SAM accessibility

```{r}
acc_SAM <- data.frame(ac(hexagons$demand,
                         eldercare$capacity,
                         distmat_km, 
                         d0 = 5,
                         power = 2, 
                         family = "SAM"))

colnames(acc_SAM) <- "accSAM"
acc_SAM <- as_tibble(acc_SAM)
hexagon_SAM <- bind_cols(hexagons, acc_SAM)
```

### 6.2 Visualising SAM’s accessibility

```{r}
tm_shape(hexagon_SAM,
         bbox = mapex) + 
  tm_polygons(fill = "accSAM", 
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
tm_shape(eldercare) +
  tm_dots(size = 0.3,
          fill = "red",
          col = "black") +
  tm_title("Accessibility to eldercare: SAM method") +
  tm_layout(frame = TRUE) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2)
```

```{r}
# Comparing all three methods in one layout

tmap_mode("plot")
# Disabling automatic resizing and accept potential overlap/clipping
# tmap_options(component.autoscale = FALSE)

tm1 <- tm_shape(hexagon_Hansen) +
  tm_polygons("accHansen", style = "quantile", n = 5, palette = "brewer.blues") +
  tm_layout(title = "Hansen Accessibility")

tm2 <- tm_shape(hexagon_KD2SFCA) +
  tm_polygons("accKD2SFCA", style = "quantile", n = 5, palette = "brewer.greens") +
  tm_layout(title = "KD2SFCA Accessibility")

tm3 <- tm_shape(hexagon_SAM) +
  tm_polygons("accSAM", style = "quantile", n = 5, palette = "brewer.purples") +
  tm_layout(title = "SAM Accessibility")

tmap_arrange(tm1, tm2, tm3, ncol = 3)
```

### 6.3 Statistical graphic visualisation

```{r}
hexagon_SAM <- st_join(hexagon_SAM, mpsz, 
                       join = st_intersects)
```

```{r}
ggplot(data=hexagon_SAM, 
       aes(y = accSAM, 
           x= REGION_N)) +
  geom_boxplot() +
  geom_point(stat="summary", 
             fun.y="mean", 
             colour ="red", 
             size=2)
```

### 6.4 Comparing correlation between accessibility measures

```{r}
# Combining all accessibility measures using original hexagons (no join)

acc_combined <- data.frame(
accHansen = hexagon_Hansen$accHansen,
accKD2SFCA = hexagon_KD2SFCA$accKD2SFCA,
accSAM = hexagon_SAM$accSAM
)

# Computing correlation matrix
acc_cor <- round(cor(acc_combined), 2)

# Visualising as heatmap
library(ggcorrplot)

ggcorrplot(acc_cor,
method = "circle",
type = "lower",
lab = TRUE,
lab_size = 4,
title = "Correlation between Accessibility Measures")
```
