---
title: "Take Home Exercise 3"
author: "Sathvika Subramanian"
date: "Nov 6, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Modelling HDB Resale Prices with Geographically Weighted Methods

# 1. Overview

This study develops a spatially informed predictive model for HDB resale prices in Singapore using transaction data from July 2024 to June 2025. Focusing on 4-room flats, the analysis integrates structural housing attributes with a rich set of geospatial accessibility measures, including proximity to hawker centres, MRT stations, parks, eldercare facilities, supermarkets, and preschool services. By converting aspatial resale records into geographically referenced point features and linking them to multiple open-source geospatial datasets, the study constructs a comprehensive analytical sandbox. Geographically Weighted Regression (GWR) is applied to uncover how price determinants vary across space, revealing localised market dynamics that conventional global models cannot capture.

# 2. Objectives

The primary objective is to build a reproducible, spatially enhanced modelling workflow capable of predicting HDB resale prices with greater accuracy by leveraging both structural characteristics and local accessibility. Specifically, the study aims to:

-   Create a reliable geospatial dataset of 4-room resale flats.
-   Derive meaningful location-based predictors through spatial distance and within-radius computations.
-   Compare global OLS and local GWR models to assess spatial heterogeneity in price determinants.
-   Communicate findings through clear, decision-oriented geovisualisations.

# 3. Installing and Loading Packages

```{r}
pacman::p_load(
  tidyverse, sf, lwgeom, stringr, lubridate,
  tmap, janitor,
  GWmodel, olsrr, performance,
  readr, here, units
)
```

```{r}
# Setting tmap plotting mode for RStudio (static plotting)
tmap_mode("plot")
options(scipen = 999)
```

# 4. HDB resale CSV

```{r}
hdb_fp <- "data/HDBresale.csv"
if(!file.exists(hdb_fp)) stop("HDBresale.csv not found at data/HDBresale.csv")

HDB <- read_csv(hdb_fp, show_col_types = FALSE) %>%
  janitor::clean_names() %>%
  mutate(
    month = as.Date(paste0(month, "-01")),
    flat_type = toupper(flat_type),
    town = toupper(str_trim(town))
  ) %>%
  filter(flat_type == "4 ROOM") %>%
  mutate(
    remaining_lease_yrs = map_dbl(remaining_lease, ~ {
      if(is.na(.x)) return(NA_real_)
      nums <- str_extract_all(.x, "\\d+")[[1]]
      if(length(nums) == 0) return(NA_real_)
      if(length(nums) == 1) return(as.numeric(nums[1]))
      as.numeric(nums[1]) + as.numeric(nums[2]) / 12
    }),
    lease_commence_date = as.numeric(lease_commence_date),
    unit_age = year(month) - lease_commence_date,
    floor_area_sqm = as.numeric(floor_area_sqm),
    storey_mid = case_when(
      str_detect(storey_range, "TO") ~ {p <- read.table(text = gsub("TO", " ", storey_range)); rowMeans(p)},
      TRUE ~ as.numeric(str_extract(storey_range, "\\d+"))
    )
  )

message("HDB rows (4-room): ", nrow(HDB))
```

# 5. Safe SF Readers Helper Functions

```{r}
safe_read_points <- function(path, target_crs = 3414) {
  if(!file.exists(path)) return(NULL)
  sf_obj <- tryCatch(st_read(path, quiet = TRUE), error = function(e) return(NULL))
  if(is.null(sf_obj) || nrow(sf_obj) == 0) return(NULL)
  gtypes <- suppressWarnings(st_geometry_type(sf_obj))
  if(!all(gtypes %in% c("POINT","MULTIPOINT"))) sf_obj <- tryCatch(st_collection_extract(sf_obj, "POINT"), error = function(e) sf_obj)
  sf_obj <- tryCatch(lwgeom::st_make_valid(sf_obj), error = function(e) sf_obj)
  sf_obj <- sf_obj[!st_is_empty(sf_obj), ]
  if(is.na(st_crs(sf_obj))) st_crs(sf_obj) <- 4326
  sf_obj <- tryCatch(st_transform(sf_obj, target_crs), error = function(e) return(NULL))
  if(nrow(sf_obj) == 0) return(NULL)
  return(sf_obj)
}

safe_read_polygons <- function(path, target_crs = 3414) {
  if(!file.exists(path)) return(NULL)
  sf_obj <- tryCatch(st_read(path, quiet = TRUE), error = function(e) return(NULL))
  if(is.null(sf_obj) || nrow(sf_obj) == 0) return(NULL)
  sf_obj <- tryCatch(lwgeom::st_make_valid(sf_obj), error = function(e) sf_obj)
  sf_obj <- sf_obj[!st_is_empty(sf_obj), ]
  if(is.na(st_crs(sf_obj))) st_crs(sf_obj) <- 4326
  sf_obj <- tryCatch(st_transform(sf_obj, target_crs), error = function(e) return(NULL))
  return(sf_obj)
}
```

# 6. Loading Supporting Layers

```{r}
mp_subzone_fp <- "data/MP14_SUBZONE_WEB_PL.shp"
hawker_fp     <- "data/HawkerCentresGEOJSON.geojson"
parks_fp      <- "data/Parks.geojson"
super_fp      <- "data/SupermarketsGEOJSON.geojson"
child_fp      <- "data/ChildCareServices.geojson"
kinder_fp     <- "data/Kindergartens.geojson"
elder_fp      <- "data/ELDERCARE.shp"

MP_subzone <- safe_read_polygons(mp_subzone_fp)
if(is.null(MP_subzone)) stop("MP14_SUBZONE_WEB_PL.shp is required but could not be read.")

hawker       <- safe_read_points(hawker_fp)
parks        <- safe_read_points(parks_fp)
supermarkets <- safe_read_points(super_fp)
childcare    <- safe_read_points(child_fp)
kinderg      <- safe_read_points(kinder_fp)
eldercare    <- safe_read_polygons(elder_fp)
if(!is.null(eldercare)) eldercare <- st_centroid(eldercare)
```

# 7. Creating HDB spatial via subzone centroids

```{r}
MP_subzone <- st_transform(MP_subzone, 3414)
subzone_points <- st_point_on_surface(MP_subzone) %>% select(SUBZONE_N, geometry) %>% st_make_valid()
subzone_lookup <- st_drop_geometry(subzone_points) %>% mutate(geometry = st_geometry(subzone_points))

HDB_geo <- HDB %>%
  mutate(town_guess = town) %>%
  left_join(subzone_lookup, by = c("town_guess" = "SUBZONE_N")) %>%
  st_as_sf(sf_column_name = "geometry", crs = 3414) %>%
  st_make_valid()

# fallback nearest subzone geometry
missing_idx <- which(is.na(st_coordinates(HDB_geo)[,1]))
if(length(missing_idx) > 0) {
  nearest_geom <- subzone_points$geometry[match(HDB_geo$town[missing_idx], subzone_points$SUBZONE_N)]
  HDB_geo$geometry[missing_idx] <- nearest_geom
}
HDB_geo <- st_make_valid(HDB_geo)
```

# 8. Jitter to avoid identical points

```{r}
# Ensuring geometry is POINT
HDB_geo <- st_cast(HDB_geo, "POINT")

# Applying jitter
set.seed(111)
coords <- st_coordinates(HDB_geo)  # now works
jitter_m <- 5
coords_j <- coords + matrix(runif(nrow(coords)*2, -jitter_m, jitter_m), ncol = 2)
HDB_geo$geometry <- st_sfc(lapply(1:nrow(coords_j), function(i) st_point(coords_j[i, ])), crs = st_crs(HDB_geo))
HDB_geo <- st_make_valid(HDB_geo)

message("Jitter applied to HDB points.")
```

# 9. Distance Functions

```{r}
nearest_distance <- function(src, dest, colname) {
  if(is.null(dest) || nrow(dest) == 0) {src[[colname]] <- 999; return(src)}
  if(st_crs(src) != st_crs(dest)) dest <- st_transform(dest, st_crs(src))
  idx <- st_nearest_feature(src, dest)
  d <- st_distance(src, dest[idx, ], by_element = TRUE)
  src[[colname]] <- as.numeric(d)/1000
  return(src)
}

within_count <- function(src, dest, dist_m) {
  if(is.null(dest) || nrow(dest) == 0) return(rep(0, nrow(src)))
  if(st_crs(src) != st_crs(dest)) dest <- st_transform(dest, st_crs(src))
  lengths(st_is_within_distance(src, dest, dist = dist_m))
}
```

# 10. Computing locational variables

```{r}
HDB_geo <- HDB_geo %>%
  mutate(resale_price = as.numeric(resale_price)) %>%
  nearest_distance(eldercare, "dist_eldercare_km") %>%
  nearest_distance(hawker, "dist_hawker_km") %>%
  nearest_distance(parks, "dist_park_km") %>%
  nearest_distance(supermarkets, "dist_supermarket_km") %>%
  mutate(
    childcare_within350 = within_count(., childcare, 350),
    kindergarten_within350 = within_count(., kinderg, 350)
  )
```

# 11. Joining subzone information

```{r}
MP_subzone <- st_transform(MP_subzone, st_crs(HDB_geo))
HDB_geo <- st_join(HDB_geo, MP_subzone %>% select(SUBZONE_N, REGION_N), join = st_intersects)
HDB_geo <- HDB_geo %>% filter(!st_is_empty(geometry) & st_is_valid(geometry))
```

# 12. Model Preparation

```{r}
predictors <- c(
  "floor_area_sqm", "storey_mid", "unit_age",
  "dist_eldercare_km", "dist_hawker_km",
  "dist_park_km", "dist_supermarket_km",
  "childcare_within350", "kindergarten_within350"
)

model_df <- HDB_geo %>% st_drop_geometry() %>%
  mutate(across(contains("dist_"), ~replace_na(., 999)),
         across(contains("within"), ~replace_na(., 0))) %>%
  select(resale_price, all_of(predictors))
```

# 13. OLS Baseline

```{r}
drop_singular_predictors <- function(df, response) {
  keep_vars <- setdiff(names(df), response)
  formula_try <- as.formula(paste(response, "~", paste(keep_vars, collapse = "+")))
  repeat {
    model <- try(lm(formula_try, data = df), silent = TRUE)
    if(inherits(model, "try-error")) break
    coefs_na <- names(coef(model))[is.na(coef(model))]
    if(length(coefs_na) == 0) break
    keep_vars <- setdiff(keep_vars, coefs_na[1])
    if(length(keep_vars) == 0) break
    formula_try <- as.formula(paste(response, "~", paste(keep_vars, collapse = "+")))
  }
  list(df = df[, c(response, keep_vars), drop = FALSE], predictors = keep_vars)
}

res <- drop_singular_predictors(model_df, "resale_price")
model_df_clean <- res$df
predictors_clean <- res$predictors

ols_formula <- as.formula(paste("resale_price ~", paste(predictors_clean, collapse = " + ")))
ols_model <- lm(ols_formula, data = model_df_clean)
saveRDS(ols_model,"data/rds/ols_model.rds")
```

# 14. GWR

```{r}
safe_scale <- function(x) { (x - mean(x, na.rm=TRUE)) / sd(x, na.rm=TRUE) }
gwr_vars <- c("floor_area_sqm", "storey_mid", "unit_age")
gwr_vars_safe <- gwr_vars[sapply(model_df_clean[gwr_vars], function(x) !all(is.na(x)|x==0))]
df_gwr_safe <- model_df_clean[, c("resale_price", gwr_vars_safe)]
df_gwr_safe <- df_gwr_safe[complete.cases(df_gwr_safe), ]
df_gwr_safe[gwr_vars_safe] <- lapply(df_gwr_safe[gwr_vars_safe], safe_scale)
keep_idx <- complete.cases(st_drop_geometry(HDB_geo)[, c("resale_price", gwr_vars_safe)])
HDB_geo_safe <- HDB_geo[keep_idx, ]
HDB_geo_safe <- HDB_geo_safe[!st_is_empty(HDB_geo_safe) & st_is_valid(HDB_geo_safe), ]

if(nrow(HDB_geo_safe)>=30 && length(gwr_vars_safe)>0) {
  HDB_sp <- as(HDB_geo_safe,"Spatial")
  gwr_bw <- bw.gwr(as.formula(paste("resale_price ~", paste(gwr_vars_safe, collapse=" + "))),
                   data = HDB_sp, approach="AICc", kernel="bisquare", adaptive=TRUE)
  gwr_fit <- gwr.basic(as.formula(paste("resale_price ~", paste(gwr_vars_safe, collapse=" + "))),
                       data = HDB_sp, bw=gwr_bw, kernel="bisquare", adaptive=TRUE)
  gwr_sf <- st_as_sf(gwr_fit$SDF)
  saveRDS(gwr_sf, "data/rds/gwr_results.rds")
} else gwr_sf <- NULL
```

# 15. Visualisation Maps

```{r}
library(tmap)

# Setting static plotting mode
tmap_mode("plot")

# Fixed plotting function for numeric SF variables
plot_numeric_sf_fixed <- function(sf_obj, var, title = NULL, size = 0.5, palette = "RdYlBu") {
  if(is.null(sf_obj) || nrow(sf_obj) == 0) return(NULL)
  if(!(var %in% names(sf_obj))) return(NULL)
  
  # Ensuring numeric
  sf_obj[[var]] <- as.numeric(sf_obj[[var]])
  
  tm_shape(sf_obj) +
    tm_dots(
      col = var,               # color by variable
      size = size,             # fixed size in map units
      palette = palette,
      auto.palette.mapping = FALSE, # disable auto-scaling
      title = var
    ) +
    tm_layout(
      legend.outside = TRUE,
      frame = FALSE
    ) +
    tm_title(title)
}

# HDB resale price map
map_price <- plot_numeric_sf_fixed(HDB_geo, "resale_price", "HDB resale price (4-room)")
if(!is.null(map_price)) print(map_price)

# GWR outputs
if(exists("gwr_sf") && !is.null(gwr_sf)) {
  
  # Identifying numeric coefficients (exclude residuals, Local_R2, X, Y)
  coef_cols <- setdiff(
    names(gwr_sf)[sapply(st_drop_geometry(gwr_sf), is.numeric)],
    c("residual", "Local_R2", "X", "Y")
  )
  
  # Plotting first coefficient as example
  if(length(coef_cols) > 0) {
    map_gwr_coef <- plot_numeric_sf_fixed(
      gwr_sf, coef_cols[1], paste("GWR coefficient:", coef_cols[1])
    )
    if(!is.null(map_gwr_coef)) print(map_gwr_coef)
  }
  
  # Plotting GWR residuals
  if("residual" %in% names(gwr_sf)) {
    map_gwr_resid <- plot_numeric_sf_fixed(gwr_sf, "residual", "GWR residuals")
    if(!is.null(map_gwr_resid)) print(map_gwr_resid)
  }
  
  # Plotting GWR Local R2
  if("Local_R2" %in% names(gwr_sf)) {
    map_gwr_r2 <- plot_numeric_sf_fixed(gwr_sf, "Local_R2", "GWR Local R2")
    if(!is.null(map_gwr_r2)) print(map_gwr_r2)
  }
}
```

# 16. Diagonstics and Outputs

```{r}
# Basic diagnostics outputs
cat("\n--- Dataset summary ---\n")
print(summary(st_drop_geometry(HDB_geo)[, c("resale_price","floor_area_sqm","unit_age","storey_mid")]))

cat("\n--- Model rows used ---\n")
print(nrow(model_df))

if(exists("ols_model")) {
  cat("\n--- OLS AIC & Adj-R2 ---\n")
  cat("AIC:", AIC(ols_model), "\n")
  cat("Adj. R-squared:", summary(ols_model)$adj.r.squared, "\n")
}

if(exists("gwr_sf") && !is.null(gwr_sf)) {
  if("Local_R2" %in% names(gwr_sf)) {
    cat("\nGWR local R2 summary:\n"); print(summary(gwr_sf$Local_R2))
  }
  if("residual" %in% names(gwr_sf)) {
    cat("\nGWR residuals summary:\n"); print(summary(gwr_sf$residual))
  }
}

message("All done. Files saved: data/rds/HDBresale_sf_integrated.rds, data/rds/ols_model.rds (if OLS fitted), data/rds/gwr_results.rds (if GWR fitted).")
```

# 17. Saving integrated dataset

```{r}
saveRDS(HDB_geo, "data/rds/HDBresale_sf_integrated.rds")
message("Pipeline completed: OLS, GWR (if enough data), and HDB_sf saved")
```
