---
title: "In-class_Ex05"
author: "Sathvika Subramanian"
date: "September 27, 2025"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# Global and local Measures of Spatial Autocorrelation

## Import libraries

```{r}
pacman::p_load(sf,sfdep,tmap,tidyverse)
```

## The Data - Importing 'Hunan' table

```{r}
hunan_sf <- st_read(dsn = "data/geospatial", layer = "Hunan")
```

```{r}
hunan_2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

## Joining Hunan and Hunan_2012

```{r}
hunan_GDPPC <- left_join(hunan_sf, hunan_2012) %>% 
  select(1:4, 7, 15)
```

## Visualising geographically weighted summary statistics

```{r}
tm_shape(hunan_GDPPC) +
  tm_fill("GDPPC",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Distribution of geographically weighted mean",
            main.title.position = "center",
            main.title.size = 1.8,
            legend.text.size = 1.2,
            legend.height = 1.50,
            legend.width = 1.50,
            frame = TRUE)
```

## Deriving Queen's contiguity weights: sfdep methods

```{r}
wm_q <- hunan_GDPPC %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(nb,
                         style = "W"),
         .before = 1)
```

## Performing Global Moran'I permutation test

```{r}
set.seed(1234)
```

```{r}
global_moran_perm(wm_q$GDPPC,
                  wm_q$nb,
                  wm_q$wt,
                  nsim = 99)
```

# Emerging Hot-Spot analysis

```{r}
pacman::p_load(sf, sfdep, tmap, 
               plotly, tidyverse, 
               Kendall)
```

## The Data

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan") %>% 
  st_transform(crs = 32650)
```

```{r}
GDPPC <- read_csv("data/aspatial/Hunan_GDPPC.csv")
```

## Creating a Time Series Cube

```{r}
GDPPC_st <- spacetime(GDPPC, hunan,
                      .loc_col = "County",
                      .time_col = "Year")
```

```{r}
is_spacetime_cube(GDPPC_st)
```

## Computing Gi*
## Deriving the spatial weights

```{r}
GDPPC_nb <- GDPPC_st %>%
  activate("geometry") %>%
  mutate(nb = include_self(
    st_contiguity(geometry)),
    wt = st_inverse_distance(nb, 
                             geometry, 
                             scale = 1,
                             alpha = 1),
    .before = 1) %>%
  set_nbs("nb") %>%
  set_wts("wt")
```

```{r}
gi_stars <- GDPPC_nb %>% 
  group_by(Year) %>% 
  mutate(gi_star = local_gstar_perm(
    GDPPC, nb, wt)) %>% 
  tidyr::unnest(gi_star)
```

## Mann-Kendall Test on Gi

```{r}
cbg <- gi_stars %>% 
  ungroup() %>% 
  filter(County == "Changsha") |> 
  select(County, Year, gi_star)
```

```{r}
ggplot(data = cbg, 
       aes(x = Year, 
           y = gi_star)) +
  geom_line() +
  theme_light()
```

## Interactive Mann-Kendall Plot

```{r}
p <- ggplot(data = cbg, 
       aes(x = Year, 
           y = gi_star)) +
  geom_line() +
  theme_light()

ggplotly(p)
```

## Printing Mann-Kendall test report

```{r}
cbg %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>% 
  tidyr::unnest_wider(mk)
```

## Mann-Kendall test data.frame

```{r}
ehsa <- gi_stars %>%
  group_by(County) %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>%
  tidyr::unnest_wider(mk)
head(ehsa)
```

## Mann-Kendall test data.frame

```{r}
emerging <- ehsa %>% 
  arrange(sl, abs(tau)) %>% 
  slice(1:10)
head(emerging)
```

## Performing Emerging Hotspot Analysis

```{r}
ehsa <- emerging_hotspot_analysis(
  x = GDPPC_st, 
  .var = "GDPPC", 
  k = 1, 
  nsim = 99
)
```

## Visualising the distribution of EHSA classes

```{r}
ggplot(data = ehsa,
       aes(x = classification)) +
  geom_bar()
```

## Visualising EHSA

```{r}
hunan_ehsa <- hunan %>%
  left_join(ehsa,
            by = join_by(County == location))
```

```{r}
ehsa_sig <- hunan_ehsa  %>%
  filter(p_value < 0.05)
tmap_mode("plot")
tm_shape(hunan_ehsa) +
  tm_polygons() +
  tm_borders(fill_alpha = 0.5) +
tm_shape(ehsa_sig) +
  tm_fill("classification") + 
  tm_borders(fill_alpha = 0.4)
```
