---
title: "In class Exercise 8"
author: "Sathvika Subramanian"
date: "Oct 25, 2025"
date-modified: "last-modified"
format: html
execute:
 echo: true
 eval: true
 warning: false
 freeze: true
---

## Getting Started

```{r}
pacman::p_load(httr, tidyverse, sf, tmap, jsonlite, progress)
```

## Geocoding HDB resale data

## Basic version

```{r}
HDBresale <- read_csv(
  "data/rawdata/HDBresale.csv") %>%
  filter(flat_type == "4 ROOM") %>%
  filter(month >= "2025-09" & month < "2025-10")
```

```{r}
HDBresale$street_name <- gsub(
  "ST\\.","SAINT",
  HDBresale$street_name)
```

```{r}
geocode <- function(block, streetname) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  address <- paste(block, streetname, sep = " ")
  query <- list("searchVal" = address, 
                "returnGeom" = "Y",
                "getAddrDetails" = "N", 
                "pageNum" = "1")
  
  res <- GET(base_url, query = query)
  restext <- content(res, as="text")
  
  output <- fromJSON(restext) %>%
    as.data.frame %>%
    select(results.LATITUDE, results.LONGITUDE)    
  return(output)
  } 
```

```{r}
HDBresale$LATITUDE <- 0
HDBresale$LONGITUDE <- 0
for (i in 1:nrow(HDBresale)){
  temp_output <- geocode(HDBresale[i,4], HDBresale[i, 5])
  
HDBresale$LATITUDE[i] <- temp_output$results.LATITUDE
HDBresale$LONGITUDE[i] <- temp_output$results.LONGITUDE
}
```

```{r}
HDBresale_sf <- HDBresale %>%
  st_as_sf(
    coords = c("LONGITUDE", "LATITUDE"),
    crs=4326) %>%
  st_transform(crs = 3414)
```

```{r}
preschool = st_read("data/rawdata/PreSchoolsLocation.kml") %>%
  st_transform(crs= 3414)
```

```{r}
within_350m <- st_is_within_distance(
  HDBresale_sf, preschool, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(prischool_350m = map_int(within_350m, length))
```


# Supplement  to Hands-on Exercise 8

```{r}
pacman::p_load(sf, spdep, GWmodel, SpatialML, 
               tmap, rsample, Metrics, tidyverse,
               knitr, kableExtra)
```

```{r}
mdata <- read_rds("data/rawdata/mdata.rds")
```

```{r}
set.seed(1234)
HDB_sample <- mdata %>%
  sample_n(1500)
```

## Checking of overlapping points

```{r}
overlapping_points <- HDB_sample %>%
  mutate(overlap = lengths(st_equals(., .)) > 1)
```

## Spatial jitter

```{r}
HDB_sample <- HDB_sample %>%
  st_jitter(amount = 5)
```

## Data Sampling

```{r}
set.seed(1234)
resale_split <- initial_split(HDB_sample, 
                              prop = 6.67/10,)
train_data <- training(resale_split)
test_data <- testing(resale_split)
```

## Multicollinearity check

```{r}
mdata_nogeo <- mdata %>%
  st_drop_geometry()
ggstatsplot::ggcorrmat(mdata_nogeo[, 2:17])
```

## Building a non-spatial multiple linear regression

```{r}
price_mlr <- lm(resale_price ~ floor_area_sqm +
                  storey_order + remaining_lease_mths +
                  PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
                  PROX_MRT + PROX_PARK + PROX_MALL + 
                  PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                  WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                  WITHIN_1KM_PRISCH,
                data=train_data)
olsrr::ols_regress(price_mlr)
```

## Multicollinearity check with VIF

```{r}
vif <- performance::check_collinearity(price_mlr)
kable(vif, 
      caption = "Variance Inflation Factor (VIF) Results") %>%
  kable_styling(font_size = 18) 

plot(vif) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Predictive Modelling with gwr

```{r}
bw_adaptive <- bw.gwr(resale_price ~ floor_area_sqm +
                  storey_order + remaining_lease_mths +
                  PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
                  PROX_MRT + PROX_PARK + PROX_MALL + 
                  PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                  WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                  WITHIN_1KM_PRISCH,
                  data=train_data,
                  approach="CV",
                  kernel="gaussian",
                  adaptive=TRUE,
                  longlat=FALSE)
bw_adaptive
```

## Model calibration

```{r}
gwr_adaptive <- gwr.basic(formula = resale_price ~
                            floor_area_sqm + storey_order +
                            remaining_lease_mths + PROX_CBD + 
                            PROX_ELDERLYCARE + PROX_HAWKER +
                            PROX_MRT + PROX_PARK + PROX_MALL + 
                            PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                            WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                            WITHIN_1KM_PRISCH,
                          data=train_data,
                          bw=bw_adaptive, 
                          kernel = 'gaussian', 
                          adaptive=TRUE,
                          longlat = FALSE)
```

## Predictive Modelling with MLR

### Predicting with test data

```{r}
gwr_bw_test_adaptive <- bw.gwr(resale_price ~ floor_area_sqm +
                  storey_order + remaining_lease_mths +
                  PROX_CBD + PROX_ELDERLYCARE + PROX_HAWKER +
                  PROX_MRT + PROX_PARK + PROX_MALL + 
                  PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                  WITHIN_350M_CHILDCARE + WITHIN_350M_BUS +
                  WITHIN_1KM_PRISCH,
                  data=test_data,
                  approach="CV",
                  kernel="gaussian",
                  adaptive=TRUE,
                  longlat=FALSE)
```

### Predicting

```{r}
gwr_pred <- gwr.predict(formula = resale_price ~
                          floor_area_sqm + storey_order +
                          remaining_lease_mths + PROX_CBD + 
                          PROX_ELDERLYCARE + PROX_HAWKER + 
                          PROX_MRT + PROX_PARK + PROX_MALL + 
                          PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
                          WITHIN_350M_CHILDCARE + WITHIN_350M_BUS + 
                          WITHIN_1KM_PRISCH, 
                        data=train_data, 
                        predictdata = test_data, 
                        bw=bw_adaptive, 
                        kernel = 'gaussian', 
                        adaptive=TRUE, 
                        longlat = FALSE)
```

## Predictive Modelling: RF method

### Data Preparation

```{r}
coords <- st_coordinates(HDB_sample)
coords_train <- st_coordinates(train_data)
coords_test <- st_coordinates(test_data)
```

```{r}
train_data_nogeom <- train_data %>%
  st_drop_geometry()
```

### Calibrating RF Model

```{r}
set.seed(1234)
rf <- ranger(resale_price ~ floor_area_sqm + storey_order + 
               remaining_lease_mths + PROX_CBD + PROX_ELDERLYCARE + 
               PROX_HAWKER + PROX_MRT + PROX_PARK + PROX_MALL + 
               PROX_SUPERMARKET + WITHIN_350M_KINDERGARTEN +
               WITHIN_350M_CHILDCARE + WITHIN_350M_BUS + 
               WITHIN_1KM_PRISCH,
             data=train_data_nogeom)
```

```{r}
rf
```